---
title: "Reproducible Programming in R Practical 3-3"
output: 
  learnr::tutorial:
    theme: lumen
runtime: shiny_prerendered
description: >
  Students run a stochastic version of the SIS model.
---

```{r setup, include=FALSE}
library(learnr)

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```

## Overview

*In this practical we develop and run our first stochastic version of the SIS model.*

## Background

We saw in the previous practicals the impact of including randomness into the
exponential model with births and deaths. Having developed the programming 
framework, it is now time to investigate the impact of randomness in the 
simplest epidemiological model – the SIS model.

## Tasks

In this practical we will write a stochastic version of the SIS model. You have 
been provided with `run_simulation()` in the RPiR package, which allows some 
extra functionality over `run_simple()` to run simulations using the older
deterministic models, and some other code you will need in later exercises. You 
need to write <span style="color: #de77ae;">0303-run-SIS.r</span> and 
<span style="color: #de77ae;">0303-stochastic-SIS.r</span>, adapting 
<span style="color: #de77ae;">0302-run-birth-death.r</span> and 
<span style="color: #de77ae;">0301-stochastic-birth-death.r</span> to handle the 
SIS model rather than exponential growth. Pay attention in particular to writing 
the function in <span style="color: #de77ae;">0303-stochastic-SIS.r</span>, 
where the most important changes need to be made.

The function will need a new name, perhaps `timestep_stochastic_SIS()`:

```{r, eval = FALSE}
timestep_stochastic_SIS <- function(latest, transmission.rate, recovery.rate, timestep) {
                                    # -- some code --
                                    }
```

It proceeds much as before with the deterministic SIS model, calculating the 
effective beta and sigma at this timestep, and then as with the stochastic birth 
death model, you need to work out whether the individual rate of infection or 
recovery per timestep is greater than 1, in which case it cannot be used as a
probability. In this case, the actual individual infection rate is 
`effective.transmission.rate` $\times \frac{I}{N}$ and the individual recovery 
rate is `effective.recovery.rate`.

Now for the key step... you need to sample from the binomial distribution (as 
described in the lecture) to determine the number of new infecteds and 
susceptibles using `rbinom()` with the probability of "success" set to the
individual rates detailed above. Then update the populations, and if there are
no infecteds left, then the epidemic is over, so set `is.finished` to `TRUE.` 

```{r, eval = FALSE}
is.finished <- next.infecteds == 0
```

Then append the new data to the data frame as usual, and return a list with the
updated population data frame and a variable that tells us whether we are 
finished or not.

```{r, eval = FALSE}
list(updated.pop = next.population, end.experiment = is.finished)
```

## Running the code / Report

When you are happy with the structure of the code, run it a few times with the 
same parameter values and observe the variability. Try changing beta (which will
change $R_0$) and the initial number of infecteds and see how the likelihood of
early extinction changes. Notice that we can automate overlaying multiple 
simulations on the same graph:

```{r, eval = FALSE}
# We want a total of 10 plots
first.graph <- TRUE 
for (loop in 1:10) {
  final.populations <- run_simulation(timestep_stochastic_SIS,
                                      initial.populations,
                                      end.time,
                                      transmission.rate = ecoli.trans, 
                                      recovery.rate = ecoli.recov, 
                                      timestep = working.timestep)
  # First time, make a new plot, afterwards just draw lines
  if (first.graph) {
    plot_populations(final.populations,
                     new.graph = TRUE, 
                     col = c(susceptibles = "green", infecteds = "red"))
    first.graph <- FALSE 
  } else {
    plot_populations(final.populations,
                     new.graph = FALSE,
                     col = c(susceptibles = "green", infecteds = "red"))
  }
}
```

You could do all ten plots in a single loop by setting `new.graph` appropriately
- by starting with `TRUE` and them changing it to `FALSE` as I did in the slides,
or changing it depending on what value `loop` has, for instance. Try changing 
the total population size and determine the impact on the observed variability. 
You should also overlay the deterministic results from Practical 2-3. 

```{r, eval = FALSE}
final.populations <- run_simulation(timestep_deterministic_SIS, 
                                    initial.populations,
                                    end.time,
                                    transmission.rate = ecoli.trans,
                                    recovery.rate = ecoli.recov, 
                                    timestep = working.timestep)
```

Write a report to demonstrate the variability in the results of the stochastic
model.
