---
title: "Reproducible Programming in R Practical 1-5"
output: 
  learnr::tutorial:
    theme: lumen
runtime: shiny_prerendered
description: >
  Students edit scripts to make a more complex model.
---

```{r setup, include=FALSE}
library(learnr)
gradethis::gradethis_setup()

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```


## Overview
*In this session we generalise the code from a growth model to a model with births and deaths.*

## Background

There are many refinements to our code that we could make, and indeed we will 
work through several of them during the course, but our current framework is 
already sufficiently sophisticated to handle some much more complex population
dynamics. We are therefore ready to start to thinking about our models rather 
than the mechanics of what we are doing.

## Tasks

In the modelling lecture we discussed a simple population growth model of the form

$$N(t+1) = N(t) + \lambda\; N(t)$$

which we can state roughly in English as 

> "the next population size is the current size plus the new additions (lambda 
times the current size)"

and this can be rewritten in turn in R as:

```{r, eval = FALSE}
next.count <- latest$count + lambda * latest$count
```

This use of greek letters is fine for mathematicians, but we use `growth.rate` 
for `lambda` in our code as we want the code to read more like English. We also 
write this on multiple lines as we calculate the new additions separately and 
then add them in. Note that, for consistency, throughout these practicals our 
code will refer to the data frame with the current latest population counts in 
as `latest` as in Practical 1-4, and these are referred to in the equations such
as those on this page as $N(t)$. The updated values, referred to above as
$N(t+1)$, appear as `next.something` in the functions if they 
have a name at all, and `updated.something` in the main script just 
to be distinct. 

We now want to update the model to include both births and deaths, *i.e.*

$$N(t+1) = N(t) + b\; N(t) – d\; N(t)$$
which can be similarly translated as

> "the next population size is the current size plus the new births (b times 
the current size) minus the new deaths (d times the current size)"

where $b$ represents the birth rate and $d$ represents the death rate.

In this practical, we will reuse 
<span style="color: #de77ae;">0104-plot-helper.r</span> while adapting 
<span style="color: #de77ae;">0104-run-growth.r</span> and 
<span style="color: #de77ae;">0104-step-growth.r</span> appropriately. For convenience, we've generated
<span style="color: #de77ae;">0105-step-birth-death.r</span> (from 
<span style="color: #de77ae;">0104-step-growth.r</span>) for you, with 
blanks for you to fill in:

```{r stepbirthdeath, exercise = TRUE, exercise.cap = "0105-step-birth-death.r", exercise.lines = 20}
#' ### Function: step_deterministic_birth_death() 
#' Run one step of a simple deterministic exponential birth and death model. 
#'
#' Arguments: 
#' - latest -- the population count now
#' - birth.rate -- the birth rate 
#' - death.rate -- the death rate 
#'
#' Returns:
#' - the updated population count
#'
step_deterministic_birth_death <- function(latest, ___, ___) {
  # Calculate changes to population
  new.additions <- ___ * latest$count 
  ___ <- ___ * latest$count 
  # Calculate population at next timestep
  next.count <- latest$count + new.additions - ___
  # Create a data frame with the updated counts and return it
  data.frame(count = next.count) 
}
```

<div id="stepbirthdeath-hint">
**Hint:** . 
</div>

```{r stepbirthdeath-solution}
#' ### Function: step_deterministic_birth_death() 
#' Run one step of a simple deterministic exponential birth and death model. 
#'
#' Arguments: 
#' - latest -- the population count now
#' - birth.rate -- the birth rate 
#' - death.rate -- the death rate 
#'
#' Returns:
#' - the updated population count
#'
step_deterministic_birth_death <- function(latest, birth.rate, death.rate) {
  # Calculate changes to population
  new.additions <- birth.rate * latest$count 
  new.subtractions <- death.rate * latest$count 
  # Calculate population at next timestep
  next.count <- latest$count + new.additions - new.subtractions
  # Create a data frame with the updated counts and return it
  data.frame(count = next.count) 
}
```

```{r stepbirthdeath-check}
grade_this_code()
```

Likewise, we've generated 
<span style="color: #de77ae;">0105-run-birth-death.r</span> (from
<span style="color: #de77ae;">0104-run-growth.r</span>), with added blanks 
for you to fill in. Again,`source()`
has been commented out as it won't work in the {learnr} environment. Instead, 
the necessary files have been preloaded for you. Try to fill in the blanks and
compare the outputs as you vary the birth and death rates:

```{r rungrowth-setup}
library(progInR)
step_deterministic_birth_death <- function(latest, birth.rate, death.rate) {
  new.additions <- birth.rate * latest$count 
  new.subtractions <- death.rate * latest$count 
  next.count <- latest$count + new.additions - new.subtractions
  data.frame(count = next.count) 
}
```

```{r rungrowth, exercise = TRUE, exercise.cap = "0105-run-birth-death.r", exercise.lines = 60}
# Load the step_simple_growth() function into the global environment (my workspace)
#source("0104-plot-helper.r") 
#source(___)

#' Set up the simulation parameters
#' --------------------------------
#' First we set up the parameters for the simulation.

# Set the growth rate
human.annual.growth <- 0.015

# Set the death rate
___ <- ___

# Starting population size
initial.count <- 7000000000

# And setting times
start.time <- 0
end.time <- 100

#' Run the simplest possible simulation
#' ------------------------------------
#' Then run it so that we can get the output we need

# Set up the population starting size (at the first timestep)
population.df <- data.frame(count = initial.count)

# The timesteps that the simulation will run through
timesteps <- seq(from = start.time + 1, to = end.time)

# Now we loop through the time itself (starting at the second timestep)
for (new.time in timesteps) {
  # Calculate population at next timestep
  updated.population <- ___(latest = tail(population.df, 1),
                            ___ = human.annual.growth,
                            ___ = human.annual.death) 
  # Add new element onto end of population vector
  population.df <- rbind(population.df, updated.population)
}

#' Plot the results
#' ----------------
#' And finally we output the results.

# Now we can plot the timesteps against the population vector
population.df$time <- c(start.time, timesteps)
plot_populations(population.df)

#' Set up a second population
#' --------------------------


#' Run the simulation
#' ------------------


#' Plot the results on the same plot
#' ---------------------------------


```

You can plot a second (or third, *etc.*) result on top of a first with commands 
like:

```{r, eval = FALSE}
plot_populations(population.b, new.graph = FALSE, col = "red")
```

Remember that `col` sets the colour of the line in the plot – use names "red", "blue", *etc.*

```{r sameanswer, echo=FALSE}
quiz(caption = "Quiz",
     question("When do you get the same answer?",
              answer("When the death rate is zero"),
              answer("When the death rate is double the birth rate"),
              answer("When the death rate is equal to the birth rate", 
                     correct = TRUE),
              answer("When the death rate plus the birth rate is zero"),
              allow_retry = TRUE,
     correct = "Correct! You should find that you get the same output if the difference between the birth and death rate is equal to the net growth rate in the previous model. This may seem a little dull and obvious but things will get more interesting when we consider stochastic models."
     ))
```
