---
title: "Practical A-1: If statements"
output:
  learnr::tutorial:
    theme: lumen
    includes:
      in_header: !expr system.file("tutorials/google-font.html", package = "RPiR")
runtime: shiny_prerendered
description: >
  Basics of control flow - if statements.
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.reveal_solution = FALSE)
gradethis::gradethis_setup()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ">",
  error = TRUE
)
set.seed(123)
```

## Control flow

### Comparisons -- testing for equality and difference

These tests are essential for `while` loops to determine
whether to continue through the next iteration of the loop, and other forms of flow control
such as `if` statements, to determine what to do next. They allow us to test whether something is true, and
change what lines of code are run depending on the outcome. There are the
basic tests:

- `a == b`: is `a` equal to `b`? (`TRUE` if this is true, `FALSE` if this is false)
- `a > b`: is `a` greater than `b`?
- `a < b`: is `a` less than `b`?
- `a >= b`: is `a` greater than or equal to `b`?
- `a <= b`: is `a` less than or equal to `b`?
- `a != b`: is `a` not equal to `b`?

There are then three basic ways of changing or combining the above:

- `(a < b) || (c == d)`: is "`a` less than `b`" **or** "`c` equal to `d`"?
- `(a < b) && (c == d)`: is "`a` less than `b`" **and** "`c` equal to `d`"?
- `!((a < b) && (c == d))`: is it **not** `TRUE` that "`a` is less than `b` 
**and** `c` is equal to `d`"?

Remember to use (lots of!) brackets to ensure you are combining things in the
right order. 

Comparisons are also [covered in passing][operators-R4DS] in
R4DS, and by R Coder in a bit more depth [here][operators-rcoder].

### `if` and `if else` statements

The `if` command allows us to perform a test, and if the result is `TRUE` run 
a block of R code (in curly brackets `{ ... }`). Optionally, if the test is
`FALSE`, a different block of code can be run instead. This allows us to do a
variety of things. To give a very simple example:

```{r ifstatement, exercise = TRUE}
if (2 > 1) {
  print("Maths works!")
}
```

This prints `"Maths works!"` because the test is `TRUE`, so the code block
that follows is run. You can try changing the numbers to get a different result. Whereas:

```{r ifstatement-2, exercise = TRUE}
if (2 >= 5) {
  print("At least 5.")
} else {
  print("Less than 5.")
}
```

This prints `"Less than 5."` because the test was false, so R
continues to the code block after the `else` statement. Again, change the numbers to check you understand how it works. If there is no `else`
statement, then no code is run, so:

```{r ifstatement-3, exercise = TRUE}
if (2 >= 5) {
  print("At least 5.")
}
```

does nothing. 

R Coder covers the basics of `if` statements [here][if-rcoder]. R4DS seems to
consider them too simple, but ironically they have [a chapter in Advanced
R][if-advanced-r2]; however, this contains a lot of advanced topics in flow
control that are well beyond the scope of this course.

[operators-R4DS]: https://r4ds.had.co.nz/transform.html?q=compa#comparisons
[if-advanced-r2]: https://adv-r.hadley.nz/control-flow.html
[operators-rcoder]: https://r-coder.com/operators-r
[if-rcoder]: https://r-coder.com/if-else-r

## Exercises

Now try it yourself. Complete the following code chunk, such that `x` is 
correctly identified as a negative number.

```{r ifexercise, exercise = TRUE}
x <- -1

if (___) {
  print("Positive number")
} else {
  print("Negative number")
}
```

```{r ifexercise-solution}
x <- -1

if (x > 0) {
  print("Positive number")
} else {
  print("Negative number")
}
```

```{r ifexercise-check}
grade_this_code()
```

Of course, this statement isn't entirely accurate, since `0` is neither positive
nor negative. You can fix this by correcting this code:

```{r ifexercise2, exercise = TRUE}
x <- 0

if (___) {
  print("Positive number")
} else if (___) {
  print("Negative number")
} else
  print("Zero")
```

```{r ifexercise2-solution}
x <- 0

if (x > 0) {
  print("Positive number")
} else if (x < 0) {
  print("Negative number")
} else
  print("Zero")
```

```{r ifexercise2-check}
grade_this_code()
```

Run the following code. There are two errors, try to find them.

```{r ifexercise3, exercise = TRUE}
x <- 3
y <- 3

if (x - y) {
  print("x is less than y")
} else if (x + y) {
  print("x is greater than y")
} else 
  print("x equals y")
```

```{r ifexercise3-solution}
x <- 3
y <- 3

if (x < y) {
  print("x is less than y")
} else if (x > y) {
  print("x is greater than y")
} else 
  print("x equals y")
```

```{r ifexercise3-check}
grade_this_code()
```

Now edit the following code chunk so that `This statement is true` is 
always returned.

```{r ifexercise4, exercise = TRUE}
if (___) {
  print("This statement is true")
} else {
  print("This statement is false")
}
```

<div id="ifexercise4-hint">
**Hint:** The value `TRUE` is always true, so will succeed in a if statement, and the value `FALSE` is always false, so will fail.
</div>

```{r ifexercise4-solution}
if (TRUE) {
  print("This statement is true")
} else {
  print("This statement is false")
}
```

```{r ifexercise4-check}
grade_this_code()
```


