---
title: "Reproducible Programming in R Practical 1-2"
output: 
  learnr::tutorial:
    theme: lumen
runtime: shiny_prerendered
description: "Reproducible Programming in R Practical 1-2!"
---

```{r setup, include=FALSE}
library(learnr)

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```


## Overview

*In this session you will take the biodiversity data package you created last week and write a demo for the package that calculates a variety of diversity measures on the data.*

## Background

Now you have a package with your data in it, we can easily run analyses, simply
starting each analysis script with the R command` library(xxx)`, where `xxx` is
the name of the package you created.

There are an enormous number of different ways of measuring diversity, to the 
extent that diversity itself has been declared a "non-concept" in the past (by
Hurlbert, 1971), but nowadays they fit broadly within Whittaker’s framework of 
three basic types of diversity: alpha, beta and gamma diversity.

* Gamma Diversity is the diversity of a metacommunity, the whole area being 
studied, whether that is an ecosystem or a single study site;
* Alpha Diversity is the average diversity of the subcommunities that together
comprise the metacommunity. The metacommunity can either be broken up into
subcommunities spatially in the case of a landscape formed of several habitats,
or temporally in the case of time series data from a single site;
* Beta Diversity is some measure of the differences between these 
subcommunities. This is the most variable in definition of the three types of
measure, including such concepts as turnover of species along a transect in
space or time, the effective number of distinct subcommunities in a 
metacommunity, the average distinctiveness of the constituent subcommunities,
and many other related concepts. Tuomisto (2010) has made a brave attempt to
make sense of these different meanings.

All of these types of diversity can be measured in several different ways. The
standard approach is often called "species diversity", where we treat all 
species as distinct and all of our measures of alpha and gamma diversity are 
counting in terms of effective numbers of species. However, we often need to 
consider similarity between species, and this results in a wide variety of
measures, from closely related ones such as counts of "distinct species"
(species that are taxonomically unrelated) or "distinct evolutionary histories"
(species that have no common ancestor in a phylogeny), to ones that are more
peripherally connected such as "distinct functional groups" (species that are
functionally distinct). Leinster and Cobbold (2012) have developed a framework 
for working with all of these types of similarity consistently. Because beta 
diversity measures are comparing subcommunities, they are generally unitless
(measuring fractions or percentages) or are measured in terms of numbers of
subcommunities; how similar two subcommunities are seen to be still entirely 
depends on the type of similarity though, resulting in, for instance a count of
"the effective number of phylogenetically / functionally / taxonomically
distinct subcommunities". The relationship between alpha, beta and gamma has 
been the source of many arguments – see, for example, Whittaker (1972), Jost
(2007) and Reeve et al. (2016).

Finally, I used the phrase "effective number" several times above. At one 
extreme, when we disregard abundance and just look at presence-absence data,
we measure "species richness" by simply counting the number of species present;
at the opposite extreme, we have measures such as Berger-Parker which only look
at dominance of the most abundant species. In between, we find measures such as
Simpson and Shannon, where we weight our counting of species by their abundance. 
For instance, Shannon entropy is a measure of the "surprise" or uncertainty 
expected from the next observation, which is low if there are few species or one
is dominant, but high when many are equally abundant. In all cases, the highest
diversity is achieved when all species are equally abundant (and completely 
distinct, when we are considering similarity). We therefore talk about the 
"**effective number of species**" as the number of *completely distinct*, 
*equally abundant species* that would give us the same value of diversity as the
observed community. This acts as a reference to allow fair comparisons to be
made. In the case of species richness, this is just the richness itself, but for
instance in the case of Shannon entropy, the *effective numbers equivalent* is 
in fact the exponential of the entropy. For diversity measures that work in 
terms of these effective numbers, the extent to which we weight by abundance is
determined by a parameter $q$, which varies from 0 – not at all, only 
presence-absence – to infinity ($Inf$) – only consider the most abundant. This 
standard for dealing with diversities as effective numbers, while far from 
universal, has now been pursued for over 40 years and is broadly accepted – 
originally by Hill (1972), and subsequently, for instance, by Jost (2007), 
Leinster and Cobbold (2012), and Reeve et al. (2016).

## Setting up R

First, you need to go back to the project that contained your diversity data by
selecting it in the top right of the RStudio window. Then make sure that all of
the correct packages are installed to do the biodiversity analyses:

```{r}
install.packages(c("vegan", "tibble", "dplyr", "magrittr", "devtools")) devtools::install_github("boydorr/rdiversity")
```

Note the second of these commands – we have updated the rdiversity package since
the last version on CRAN. `install_github()` installs the latest development 
version of a package (if it is developed on the github website, as many are)
before it is uploaded to CRAN. You will also need to make sure your package 
depends on these other packages, so that they are automatically installed when
you install it. To do this you need to add them to the imports section of the
`Imports:` section of the **DESCRIPTION** file. Any package that you use in your
script (i.e. you call (`library(pkgname)` somewhere) should go into the
**DESCRIPTION** file.

Now you need to create a demo inside the package. First go to the 
<span style="color: #1697C0;">Files</span> tab and create a new folder called
**demo**. You can find details on creating a demo in 
http://r-pkgs.had.co.nz/demo.html but essentially you just need to create an R
script **with a name with no spaces in it** in that folder, and then create a
new text file also in that folder called **00Index**. Each demo you create is 
entered in one line of this file with the demo name followed by a one-line
description. For example, the demo file **diversity-of-BCI.R** would have a line 
like:

```
diversity-of-BCI Showing the diversity of the BCI forest plot.
```

More importantly, however, you need to create the R script that will be the 
demo, and save it in the **demo** folder. Note that demos should never contain 
any commands to install packages, the above code is not part of the demo! The
script will instead be measuring several kinds of alpha, beta and gamma
diversity, detailed below, but in order to do some work with measures of
taxonomic diversity of the BCI dataset, the taxonomy will first need to match
the species abundance data exactly. Here I’m assuming that you created the 
package `{my_package}`, and in it `bci_taxa` contains the taxonomy of all BCI 
species and `bci_2010` contains the 2010 BCI census:

```{r}
library(my_package)
library(dplyr)
# First filter for the correct species, either using dplyr: 
taxonomy <- bci_taxa %>%
  filter(GenusSpecies %in% rownames(bci_2010)) %>%
  unique() %>% as.data.frame()
# or using convential code:
taxonomy <- as.data.frame(unique(bci_taxa[bci_taxa$GenusSpecies %in% rownames(bci_2010),]))
# Then order the species correctly:
rownames(taxonomy) <- taxonomy$GenusSpecies 
taxonomy <- taxonomy[rownames(bci_2010),]
```

To create a metacommunity to analyse in the rdiversity package, you will need to
create an object of that type using `metacommunity()`, and then you can 
calculate, for example, the gamma diversity of the whole metacommunity with the
command `meta_gamma()` and plot the output with `plot()`:

```{r}
library(rdiversity)
meta <- metacommunity(bci_2010)
results <- meta_gamma(meta, qs = seq(from = 0, to = 5)) 
plot(diversity ~ q, type = "l", data = results)
```

To create a metacommunity that takes account of similarity between species, you
need to input it as a second argument to `metacommunity()`, using two functions
– one, `tax2dist()`, to create a distance matrix from a taxonomy, and a second,
`dist2sim()`, to create a similarity matrix from the latter. So:

```{r}
taxSim <- tax2dist(taxonomy, c(GenusSpecies=0, Genus=1, Family=2, Other=3)) %>%
  dist2sim("linear") 
metatax <- metacommunity(bci_2010, taxSim)
results4tax <- meta_gamma(metatax, qs = seq(from = 0, to = 5)) 
lines(diversity ~ q, col=2, data = results4tax)
```

Finally, to create a metacommunity to analyse in vegan, you need to transpose the axes of the abundance data matrix. So:

```{r}
library(vegan)
abundances <- t(bci_2010) # this swaps rows and columns specnumber(abundances,groups=1) # Calculate species richness (gamma)
```

loads the dataset and then calculates the species richness of the whole
metacommunity. **It’s important to always use the `abundances` variable when using the vegan package, and always use the `bci_2010` variable when using rdiversity – they need the data in opposite formats!**

## Tasks

First, read the documentation and examples at 
https://github.com/boydorr/rdiversity and/or in R. Then write a script to 
calculate and plot the gamma diversity and the normalised alpha diversity for 
the 2010 BCI census, for a range of values of $q$, say `qs = 0:5`. Then repeat
using taxonomic similarity, showing that there are fewer taxonomically distinct
species than there are species. Now do this again counting the effective number 
of genera and families. Note that this can be achieved by using `tax2dist()` 
with `c(GenusSpecies=0, Genus=0, Other=1)` and 
`c(GenusSpecies=0, Genus=0, Family=0, Other=1)`, respectively – these distance
measurements say that two species in the same genus (or family) are zero
distance apart, just like two individuals in the same species. As a result, the
diversity measures following these instructions will simply count in effective
numbers of genera or families, respectively, rather than species.

Moving on to beta diversity, we now want to find out which quadrats in the BCI 
study site are the most distinct. We can measure average distinctiveness of all 
of the quadrats in the plot using `raw_meta_beta()`, but you can also ask for 
the distinctiveness of every quadrat individually using `raw_sub_beta()`. 
Calculate that for a single value of q, say qs=1, and show a histogram of the
distinctivenesses (the diversity column) of the subcommunities. Sort the data 
frame by distinctiveness and output just the highest values. Bearing in mind 
that the first two digits of the quadrat name are an x-coordinate, and the
second two are a y-coordinate, is there any pattern about these distinct areas?
If so write about it in the report.

A similar option in vegan is the function `betadiver()`. Look at the 
documentation in R for this function – you will see that it can use a very large
range of different beta diversity measures – and calculate the beta diversities
for the subcommunities using the `"j"` Jaccard and `"sor"` Sorensen similarity
measures. This function calculates pairwise similarities between the quadrats as
`dist` objects. You can convert these to a matrix and then calculate the average
similarities to all of the quadrats by using `as.matrix()` and then 
`rowMeans()`. You should be able to show that some of the identified quadrats 
(with the lowest average similarity to the other quadrats) are the same, even
though the exact beta diversity measure, and indeed even its direction, is 
different.

Remember if you want to add some plain text into a report, you just write it 
after a `#'` comment:

```{r}
#' We can make an observation here that Sorensen, Jaccard and
#' rdiversity's beta diversity measures identify overlapping
#' quadrats as the least similar / most distinct and it will appear #' as plain text in the report.
```

Generate a report from the script(s) – for each file go to the 
<span style="color: #1697C0;">File</span> menu, choose 
<span style="color: #1697C0;">Compile Report...</span> and choose the 
<span style="color: #1697C0;">html</span> output format and press 
<span style="color: #1697C0;">Compile</span>. A webpage should pop up in a 
viewer after a couple of seconds, and a new file should appear in the directory 
you are working in. You can check that this has really worked by opening the 
file in a web browser. At this point, you can improve the commenting on the 
document, so that it uses more of the features of the markdown format. Once 
you’re happy with it, generate a notebook again, and check it seems okay. Now
reinstall the package, call `library(BCI)` and run `demo("diversity-of-BCI")` or
whatever you called the package and script – make sure that it runs as expected 
when you do this. Alternatively you can call 
`demo("diversity-of-BCI", package="BCI")` without loading the library first.

## Reporting the results

You now have a demo file that will install with your package and show examples
of diversity calculations of the BCI data. You should now commit all of these 
changes (except the html file generated as the report, you should delete this)
using git and push the committed changes to GitHub. Now open an issue on GitHub 
asking your partner/collaborator to test the package for you.

Your collaborator should now check the package to make sure that they can see
and run the demo successfully. As the collaborator, you should attempt to 
install the package again using `devtools::install_github()`. You can then test
the demo by using the `demo()` function. You can also generate the report from
the demo as follows:

```{r}
rmarkdown::render(system.file("demo/diversity-of-BCI.R", package="BCI"), output_dir = ".")
```

The report will be saved to the base folder of your current project (or whatever 
your current working directory is). Check that it looks okay by opening it in a
web browser (you can delete it afterwards). If everything looks okay, or if you
identify a problem, you should reply to the issue saying what you find. 
Depending on whether there was a problem, you may need to go back and forth, or
that may be everything finished...
