---
title: "Project 1: Species richness and Simpson diversity"
output:
  learnr::tutorial:
    theme: lumen
    includes:
      in_header: !expr system.file("tutorials/google-font.html", package = "RPiR")
runtime: shiny_prerendered
description: >
  Final project, creating species richness and Simpson diversity functions
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.reveal_solution = FALSE)
gradethis::gradethis_setup()

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```

## Overview

In this series of practicals you will develop the code to calculate a few
diversity measures. In this first exercise we will calculate the species
richness and Simpson diversity of a population.

### Background

R provides us with many built-in datasets, but sadly none of them are counts of
species abundances, so you will use the famous Barro Colorado Island dataset, a 
30+ year study of a 50-hectare forest plot in Panama. We will use the number of 
trees in the study site in 2010, which you have access to from the package you
developed in Practical Series 4.

You will also construct a couple of fake datasets that you will use for 
comparison. Calculating the diversity of these populations (remembering that 
there are many, in fact a continuum of, diversity measures) will tell us how 
many observed species there are -- the species richness -- and give us measures 
of how evenly distributed the trees are amongst the tree species (the evenness 
of the species distribution) by various measures. **Species richness** is a 
very commonly used measure of diversity, which simply **counts the number of
non-zero species abundances**; **Simpson diversity** is also commonly used: it
simply **calculates the species proportions from the counts (so the total sums 
to 1), squares them all individually, and add them all together** -- it tells us
(amongst other things) how likely two randomly chosen individuals are to be 
from the same species.

## Tasks

This project will be written in the **githubusernameDiversity** package that
you set up in the last exercise, with functions being saved in the
<span style="color: #1697C0;">R</span> folder and documented so that they work
with the standard R help system, and scripts being saved as demos in the
<span style="color: #1697C0;">demo</span> folder so that they can be run using
the `demo()` function or compiled into reports to read the results with suitable
reporting explaining what is happening. In each project exercise you will have
to write some functions and (most likely) one script. *So now create a demo
script for this project and add an entry for it in the
<span style="color: #1697C0;">00Index</span> file in the same folder.*

### The study populations

We are taking the data from the 2010 census -- the true abundances for the whole 
site can be calculated as follows:

```{r, eval = FALSE}
library(githubusernameBCI) 
tree.pop <- rowSums(bci_2010)
```

The total counts of trees of different species in this forest plot are now 
available in the variable `tree.pop`. We can then do some simple extraction of
information from the full species counts:

```{r, eval = FALSE}
species.names <- names(tree.pop) 
num.sp <- length(tree.pop) 
num.trees <- sum(tree.pop) 
mean.trees <- mean(tree.pop)
```

The next two lines sample a single random quadrat and sum up 10 random 
quadrats, and store the results in `quadrat.pop` and `quadrat10.pop`,
respectively:

```{r, eval = FALSE}
quadrat.pop <- rowSums(sample_by_subcommunities(bci_2010, 1)) 
quadrat10.pop <- rowSums(sample_by_subcommunities(bci_2010, 10))
```

You will also use a handful of fake datasets for comparison:

```{r, eval = FALSE}
one.pop <- c(num.trees, rep(0, num.sp - 1)) 
uneven.pop <- c(num.trees / 2, rep(0, num.sp - 1)) + rep(mean.trees / 2, num.sp) 
mixed.pop <- (1:num.sp) / sum(1:num.sp) * num.trees
even.pop <- rep(mean.trees, num.sp)
rand.pop <- as.vector(rmultinom(1, num.trees, rep(1 / num.sp, num.sp)))
rand50.pop <- as.vector(rmultinom(1, round(num.trees / 50), rep(1 / num.sp, num.sp)))
```

Remember that `rep(x, size)` repeats `x`, `size` times to make a vector of
length `size`, and `rmultinom()` draws from the **multinomial distribution**, 
in this case, `rand.pop` is a population of 222,715 individuals 
drawn from 299 species, each equally abundant ($\frac{1}{299}$). Note that
`rmultinom()` is is in the `stats` package, so you need to add this to the
dependencies of your package now. We won't tell you about any future
dependencies you need to add, but note that you can check what package a
function is in by looking in the help file for the function:

```{R, eval = FALSE}
?rmultinom
```

If you run this in the console, the **Multinom {stats}** in the top left of the
help page tells you that this function is in the `stats` package.

### A species richness function

Then write and document a function that takes a vector of abundances like the
data above as an argument and calculates the species richness from it. Remember
that the species richness is the number of species whose abundance is not zero.
The name of this function should start with the letters `p1_`, so if you decide
to call it `species_richness()`, you should actually call it
`p1_species_richness()`. This is because you will be making a series of similar
functions throughout the project, so they will need to be called
`p1_species_richness()`, `p2_species_richness()`, `p3_species_richness()`,
*etc.* to distinguish them.

*Hint:* Applying a test to a vector 
of numbers, will apply the test to all of the numbers in the vector and produce 
a vector of booleans (`TRUE` or `FALSE`) values, so:

```{R, comment = ""}
c(1, 3, 5, 7) > 4
```
1 and 3 are not `> 4` and 5 and 7 are. And then `sum()` will count the number of
`TRUE`s in such a vector of booleans. So:

```{R, comment = ""}
over4s <- (c(1, 3, 5, 7) > 4)
sum(over4s)
```

If you think about it, you should be able to see how to use these two ideas to
count the number of non-zero abundances in a population.

### A Simpson diversity function

Then write and document a second function that will calculate the Simpson 
diversity of the population, remembering, first, that this measure works with 
the relative proportions of the species not the total counts and, second, that 
we can do calculations with the whole vector at once.

### Running the code

Look at the values in the nine population vectors -- you could try sorting them 
using `sort(x.pop)` and plotting them using `plot(sort(x.pop))`, 
`barplot(sort(x.pop))` (or something more attractive?) to get a clearer idea of 
what they look like -- and decide which you think have the richest species 
diversity.

Run the two functions on the populations provided as see whether the results of 
the two diversity measures agree with your assessments. Note that a lower value 
means a higher diversity for the Simpson diversity measure.

**NB** Remember that to make the functions you have written accessible to
your demo, you need to document and install them:

```{R, eval = FALSE}
devtools::document()
devtools::install()
library(githubusernameDiversity)
?p1_species_richness
```

You also need to install the package to use the demos:

```{R, eval = FALSE}
devtools::install()
demo(package = "githubusernameDiversity")
```

## Demo

Edit the demo that you have started above so that it loads in the functions (by
calling `library(githubusernameDiversity)`),  reads in the BCI data, and
generates a report that shows some kind of plot of each of the series of counts
and then calculates their species richness and Simpson diversity. Check that
your functions also generate appropriate help files
(*e.g.* `?p1_species_richness`).

Remember to commit the code once you are happy before moving on to the next 
exercise -- you can always come back later and make further commits if you 
realise you need to change something. I won't remind you about this in 
subsequent exercises.

### Note on function names

As we said above, note that you will often need to refer back to functions in
later exercises (as you did in the practicals) and you will be writing several
very  similar functions in successive exercises. You will need to make sure they
do not have identical names or you won't be able to call them all. 
The simplest way of doing this might be to label them by the project exercise
number and call them `p1_species_richness`, `p2_species_richness`, *etc.* 
depending on what exercise you are in, but you are welcome to find a more 
elegant solution -- anything that distinguishes them is fine.
