---
title: "1-3:Exponential growth model"
output:
  learnr::tutorial:
    theme: lumen
    includes:
      in_header: !eval.expr system.file("tutorials/google-font.html", package = "RPiR")
runtime: shiny_prerendered
description: >
  Students learn how to separate functions from the main script and source() 
  them.
---

```{r setup, include=FALSE}
library(learnr)
gradethis::gradethis_setup()

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```

## Overview
*In this session you will make it easier to reuse your code by placing functions in separate files, so that they can be referred to directly without having to be copied and pasted from file to file – a potentially error-strewn pastime.*

## Background

Breaking down the program into separate files is an important habit to get into. 
It makes the code less prone to error in the longer term and easier to debug 
because by creating independent units we are creating generic pieces of code 
that we can re-use over and over again in other programs. Not only is reusing 
code by copying and pasting code from one file to another inherently dangerous 
as you may accidentally copy the wrong piece of code, but if a mistake is 
identified at a later date there may be several places where the mistake has been 
copied to and you may not remember them all, causing avoidable mistakes further 
down the line.

## Tasks

We can turn the single program from the previous practical into two pieces of 
code by putting the `step_simple_growth()` function into a separate file, and 
reading that file in the original script using the `source()` function. We have
called the new file containing the function 
<span style="color: #de77ae;">0103-step-simple-growth-function.r</span>, so we 
will use the command:

```{r, eval = FALSE}
source("0103-step-simple-growth-function.r")
``` 

to tell the main script (the R program) that it needs to look in 
<span style="color: #de77ae;">0103-step-simple-growth-function.r</span> for
code. 

The function documentation should be in the function file now and not 
in the main script, like this:

```{r functions, exercise = TRUE, exercise.cap = "0103-step-simple-growth-function.r", exercise.lines = 18}
#' ### Function: step_simple_growth() 
#' Run one step of a simple deterministic exponential growth model. 
#'
#' Arguments: 
#' - current.population -- the population count now
#' - growth.rate -- the growth rate 
#'
#' Returns:
#' - the updated population count
#'
step_simple_growth <- function(current.population, growth.rate) {
  # Calculate changes to population
  new.additions <- growth.rate * current.population
  # Calculate population at next timestep
  next.population <- current.population + new.additions
  # Return updated population
  next.population
}
```

and the main script (<span style="color: #de77ae;">0103-run-simple-growth.r</span>) 
should look 
something like this:

> Note that {learnr} doesn't create files, so `source` doesn't work here. To 
avoid an error, we commented out line 2, and instead preloaded
<span style="color: #de77ae;">0103-step-simple-growth-function.r</span> into the
following code block)

```{r main-setup}
step_simple_growth <- function(current.population, growth.rate) {
  new.additions <- growth.rate * current.population
  next.population <- current.population + new.additions
  next.population
}
```

```{r main, exercise = TRUE, exercise.cap = "0103-run-simple-growth.r", exercise.lines = 42}
# Load the step_simple_growth() function into the global environment (my workspace)
#source("0103-step-simple-growth-function.r")

#' Set up the simulation parameters
#' --------------------------------
#' First we set up the parameters for the simulation.

# Set the growth rate
human.annual.growth <- 0.015

# Starting population size
initial.count <- 7000000000

# And setting times
start.time <- 0
end.time <- 100

#' Run the simplest possible simulation
#' ------------------------------------
#' Then run it so that we can get the output we need

# Set up the population starting size (at the first timestep)
population.vector <- initial.count

# The timesteps that the simulation will run through
timesteps <- seq(from = start.time + 1, to = end.time)

# Now we loop through the time itself (starting at the second timestep)
for (new.time in timesteps) {
  # Calculate population at next timestep
  updated.human.population <- step_simple_growth(current.population = tail(population.vector, 1), 
                                                 growth.rate = human.annual.growth)
  # Add new element onto end of population vector
  population.vector <- c(population.vector, updated.human.population)
}

#' Plot the results
#' ----------------
#' And finally we output the results.

# Now we can plot the timesteps against the population vector
plot(c(start.time, timesteps), population.vector, type = "l")
```

When we run the code, we don’t even need to open the file 
<span style="color: #de77ae;">0103-step-simple-growth-function.r</span>, 
we just run the main script
(<span style="color: #de77ae;">0103-run-simple-growth.r</span>) making sure that 
we have specified 
<span style="color: #de77ae;">0103-step-simple-growth-function.r</span> as a 
`source()` in it. Its contents (the `step_simple_growth()` function) are safely 
packaged away unless we want to edit them. 
<span style="color: #de77ae;">0103-run-simple-growth.r</span> and 
<span style="color: #de77ae;">0103-step-simple-growth-function.r</span>
together do **exactly** what 
<span style="color: #de77ae;">0102-growth-functional.r</span> did. 
