---
title: "Reproducible Programming in R Practical 1-3"
output: 
  learnr::tutorial:
    theme: lumen
runtime: shiny_prerendered
description: "Reproducible Programming in R Practical 1-3!"
---

```{r setup, include=FALSE}
library(learnr)

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```


## Overview
*In this session you will add some functions to your biodiversity data package to allow you to subsample from your data. The intention is just to see how we add functions to a package, and to provide you with an example of what the documentation for a function in an R package looks like. You will also add the documentation for the package as a whole.*

## Background

It’s rarely the case that you will have access to a fully sampled dataset like
that at BCI. To pretend that your dataset has incomplete data, I have provided 
you with two functions on Moodle that will subsample the data in different ways. 
The first – `sample_by_species()` – will sample from the dataset as if you were
recording only some of the species (chosen at random) and ignoring the others. 
The second – `sample_by_subcommunities()` – will record the counts from only 
some of the subcommunities (here quadrats) and ignore the others. There are 
other ways of subsampling or recording data incompletely, and we will
investigate this in the project. You can see some documentation on how to write
help files for functions at https://r-pkgs.org/man.html#man-functions and an
older version at http://r-pkgs.had.co.nz/man.html#man-functions – these pages 
are very useful and you should take a look.

## Tasks

You simply need to download the R files for these functions from Moodle, put
them in the **R** folder of your package, edit the documentation for each of 
these functions, and the **DESCRIPTION** file of your package so that all 
dependencies are included. Then regenerate the documentation and reinstall the
package from the Build menu. You should now be able to access the help files for
these functions (`?sample_by_species` and `?sample_by_communities`) and run the
example code (by calling `example(sample_by_species)`, for instance). Finally,
you should create a description for the overall package, so that ?BCI (in my
case) will return a description, and you should put the same (or similar text
into the README.md file (using Git Flavored Markdown) for the package so that
people going to GitHub will see what the package that you have created does
without having to install it.

This is done in the following way. Create a new file in the **R** folder – I’m 
going to call it **BCI-package.R**. Then describe your package (most easily
lifted from the **DESCRIPTION** file):

```{r}
#' BCI data package #'
#' Package to hold the BCI data (or whatever) – maybe also mention
#' something about these functions now, and put that in the
#' DESCRIPTION too. And then put it in the README.md file. #'
#' @import magrittr #'
#' @name BCI-package
#' @aliases BCI
#' @docType package #'
NULL
```

There are a few things going on here that you should notice for making packages 
in the future. The first is that you need to say what your package is called – 
here I have given it two names `BCI-package` and an alias of just `BCI`, which 
you do with the `@aliases` command. Secondly, the `NULL` to the end of the file
is included when there is no object associated with this documentation, which is
the case here, since this file contains the package documentation, as defined
explicitly in the `@docType` tag (more info here –
http://r-pkgs.had.co.nz/man.html#man-packages). Finally, if you are using any 
packages in any of your functions, you may want to import them into your 
package. This is done using `@import` – see more here –
http://r-pkgs.had.co.nz/namespace.html. You can then put the same (or similar)
text into the **README.md** file (documentation here –
https://guides.github.com/features/mastering-markdown/), but note with markdown
you don’t need to start lines with `#'`. Otherwise things are very similar. Once
you have redocumented and reinstalled everything, and checked that it is all
working, commit and push all of the changes to GitHub and ask your collaborator
to check that this works for them as well.
