---
title: "Reproducible Programming in R Practical 2-4"
output:
  learnr::tutorial:
    theme: lumen
runtime: shiny_prerendered
description: "Reproducible Programming in R Practical 2-4!"
---

```{r setup, include=FALSE}
library(learnr)
gradethis::gradethis_setup()

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```

## Overview

*In this session we will start to think ahead about what kind of framework we will need when we move on to more complex models. We will build this framework (using data frames) in the context of our existing exponential growth model, and we will separate out another potentially complex element of the code – the plotting function – where we can refine it separately from the functionally important simulation code.*

## Background

Our current model of growth of a population is held in the function 
`step_simple_growth()`, and has only one parameter (`growth.rate`), one group of 
individuals (`current.population`), and you could easily consider the amount of 
code and commenting that has gone into updating the latter with the former 
overkill.

However, we are developing this code knowing that we are going to be moving on 
to significantly more complex models with multiple parameters (*e.g.* infection 
rate, recovery rate), and multiple groups (*e.g.* susceptible and infected 
individuals), more complex ways of updating the population structure (of which 
more later), and more sophisticated ways of displaying the results.

With that in mind we are going to see how to make some of the structures in the 
code more general by introducing **data.frames**, which are just R’s way of 
introducing a table of data. Although the advantages of this are somewhat 
limited at the moment, they will be hopefully be clear when we move on to the 
epidemiological models.

## Tasks

The task for this practical is to replace the vector `population.vector` with a 
**data frame**, which we are calling `population.df`, with just one column which 
we have called `count`. We want to end up with a population update function 
(which we have now called `step_deterministic_growth()` for consistency with 
later programs) that takes in a data frame containing the current population 
count and the `growth.rate` parameter and outputs a data frame containing the 
updated count. Rewrite the code throughout so that it uses a data frame instead 
of a vector – we have called these files **0204-run-growth.r** and 
**0204-step-growth.r**, but obviously exact namings are up to you. Remember to 
use `data.frame()` to create a data frame and `rbind()` to append one onto 
another. (See the help section on the next page for step-by-step details on the 
changes).

So that we don’t need to worry about how to plot the outputs, we have provided a 
function called `plot_populations()` in **0204-plot-helper.r** which will take a 
data frame with any number of population columns and a single corresponding 
`time` column and plot them together in different colours with a legend 
indicating which line corresponds to which population. The file 
**0204-plot-helper.r** contains the function, so you will need to `source()` 
this file in your code. In order to use it you will need to add a column called 
`time` into your data frame with the times in it and then call:

```{r}
plot_populations(population.df)
```

## Help

We are now breaking the code into three sections **0204-run-growth.r**, 
**0204-step-growth.r** and **0204-plot-helper.r** which handle the mechanics, 
the model and the display respectively. The last of these is provided by us. **We will talk about how to change the function later, but first we’ll work on the main script. Note that the script won’t run till you have updated the function, so you may prefer to do it in the opposite order – just go to the next page if so.**

Changes needed to update **0203-run-simple-growth.r** to **0204-run-growth.r**

**0204-run-growth.r** will take the place of to **0203-run-simple-growth.r** in
the previous practical, and differs in the following ways. At the beginning of 
the file, it should refer to functions in the other two, separate, files:

```{r}
source("0204-plot-helper.r") 
source("0204-step-growth.r")
```

The second, and biggest change throughout **0204-run-growth.r** (and 
**0204-step-growth.r**) is that the population is now defined in terms of a data 
frame rather than as a vector. So setting it up will happen like this:

```{r}
population.df <- data.frame(count = initial.count)
```

A data frame is just R’s ways of referring to a table of data, so whereas our 
vector approach in the previous code would allow us to have a single list of 
values, the data frame will allow us to have several columns of data. Whilst 
this isn’t necessary for this piece of code the approach will be very useful 
when we are handling multiple populations (for example susceptible, infected and
recovered individuals in an epidemiological model).

Now we want to send the latest row of the data frame to the function 
`step_deterministic_growth()`, which we will define below. It turns out we do 
that exactly the same way (using `tail()`) as we did with vectors. Notice that 
we have changed the name of the first parameter to something generic which will 
work in all future models even if they are not about population dynamics.

```{r}
updated.population <- step_deterministic_growth(latest = tail(population.df, 1),
                                                growth.rate = human.annual.growth) 
population.df <- rbind(population.df, updated.population)
```

The second line is slightly different, however. `rbind()` binds – appends – the 
rows of a second data frame after the end of the last row of the first data 
frame. The two data frames must have the same columns for this to work, or they
wouldn’t fit: in this case they will both have to have one column called 
`count`, so our function will need to return a data frame with just that column 
in it when we write it below. You will get a clearer idea of how the data frames 
work once we start on the epidemiological models with multiple populations, so 
don’t worry if it isn’t currently clear.

At the end of your new **0204-run-growth.r** file, rather than use R’s basic 
plot command plot as we did previously, we have now created a function `plot_populations()`, which is defined inside **0204-plot-helper.r**. The job of 
this function is to take a data frame with any number of population columns and 
a single corresponding `time` column and plot them together in different colours 
with a legend indicating which line corresponds to which population. It will
allow us to ignore the plotting details for the moment.

In order to use it you will first need to delete the current `plot()` command at 
the end of your new file (if you copied it in), and instead add a column onto 
your data frame with the times in and use `plot_populations()` on this updated 
data frame. As the `timesteps` vector is missing the start time of the 
simulation, so the column you need to add to the data frame is actually
`c(start.time, timesteps)`.

```{r}
population.df$time <- c(start.time, timesteps)
```

Then you will call:

```{r}
plot_populations(population.df)
```

**Changes needed to update 0203-step-simple-growth.r to 0204-step-growth.r**

To match the change of population to a data frame, we now adjust **0203-step-simple-growth.r** accordingly. **0204-step-growth.r** contains the function doing the work in the simulation, now called `step_deterministic_growth()` – but with a few improvements to work in terms of a data frame rather than just a vector. Let’s go through the components of `step_deterministic_growth()`

```{r}
step_deterministic_growth <- function(latest, growth.rate) {
  ## Calculate population changes
  new.additions <- growth.rate * latest$count 
  next.count <- latest$count + new.additions
  ## Create a data frame with the updated counts and return it
  data.frame(count=next.count) 
  }
```  

We are now allowing the populations that are handled within this function to 
potentially contain several columns corresponding to different sets of 
individuals. Thus, `latest` is a data frame that contains a single row
containing the latest population size for each of the component populations – in 
this case just the one `count.` To perform the update step we need to extract 
the actual population we are interested in. For this we use the `$` expression.

**Tip:** *In general if we have a data frame called `my.frame` then `my.frame$first.col` will give us the column called `first.col`. So in our example (although we only have one column) we want to extract the column called count from the data frame latest with the expression `latest$count` and use its value for the update step.*

Since names of variables inside functions are irrelevant, we have decided not to even give one to the data frame we return. The function will just return the data frame, and it will be assigned to whatever variable in the main script we choose, in this case updated.population.

## Running the code

Again, the code should produce the same output as the previous three practicals, 
but perhaps with a slightly prettier and clearer display. Try, for instance,
generating a second result with different initial parameters and then plot the
results using:

```{r}
plot_populations(population.df, new.graph = FALSE, col = 2)
```

**Note:** we do not expect you to be able to write a function like this right 
now, but you do have experience with writing R graphics code already, so feel 
free to edit it if you really want to – you can even play around with the 
arguments available to the function (see the comments in the file for details) 
and try running the commands used. The extra argument `col` sets the colour of 
the line in the plot – either use numbers (1 is black, 2, 3, etc.) or names 
"red", "blue", etc.; or you could use `lty` to set the type of line – solid, 
dash, dot, etc. – by using numbers (1 is solid, 2, 3, *etc.*). Remember than you 
can ask for help on built-in commands (like `legend()`) by typing in the R 
console:

```{r}
?legend
```

For a very few commands you cannot ask for help using the `?` syntax – `?for`, `?function` and a couple of others. In that case you have to go to the help tab
in RStudio and type the name into the box in the top right. Note that
`plot_populations()` also has a help file if you install it using
`devtools::install_github("boydorr/ProgInR")` as we showed you earlier, and then
accessed using `library(ProgInR)`. If you just `source()` it (as we suggest)
then the only documentation is that in the file itself.

As far as the `plot_populations()` function is concerned, feel free to refer 
back to this code later to understand better how it works – there’s also a 
simpler version of this function in the same file, which provides more limited functionality but you may find easier to read, called `plot_simple()` – but some
of the helper functions we provide later in the course are fairly sophisticated 
and you will probably have to take on trust for the time being. We will be
providing these functions as we go along to allow you to focus on the models 
being developed – you are welcome to use them yourselves later, but if you use 
them for real you should spend some time satisfying yourself that they really do 
what you want.

## Report

Again, generate a notebook from the appropriate files as your report and check 
they are still generating the right results, though now with a marginally
prettier figure! Feel free to adapt `plot_populations()` to do more advanced 
things if you like – *e.g.* to move the legend about – but I don’t mind at all
if you just use the function as is, with all of its existing limitations.
