---
title: "Reproducible Programming in R Practical 2-3"
output: 
  learnr::tutorial:
    theme: lumen
runtime: shiny_prerendered
description: >
  Students improve their timestep function. They also learn how to use while() 
  loops.
---

```{r setup, include=FALSE}
library(learnr)

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```

## Overview

*In this practical we separate out the model from the simulation functionality more cleanly. In the process we make the previous set of code safer to use (and more elegant!).*

## Background

When we are programming we often make quick additions to code, as we did in the
last practical but it is good practice, if we want to keep them, to make sure
they are included in the code in a way that makes it clear what’s going on. We
had to know how the SIS model works to know that we could get away with 
multiplying the parameters with `timestep.` This use of a low-level detail of
how the model runs might be risky – we may later change the model and 
not realise this hack no longer works.

## Tasks

In the previous practical we multiplied the transmission and recovery rates up
(or down) to reflect changes in `timestep.` However, `transmission.rate` and
`recovery.rate` were not really changing, and we want the code to reflect that
rather than abusing it to do something it wasn’t intended to do. Instead add a
new argument, `timestep`, into `step_deterministic_SIS()` – we could call it
`timestep_deterministic_SIS()`. Rather than submit the product 
`transmission.rate * timestep` as an argument to the simulation function, we can 
now separately pass the real rate parameters and the `timestep` separately. This **rescaling** is now done within the function
`timestep_deterministic_SIS()` in <span style="color: #de77ae;">0203-deterministic-SIS.r</span>. We might, for
instance, create two new variables, perhaps called `effective.transmission.rate` 
and `effective.recovery.rate` (or something shorter!). These new variables are 
just our original rates rescaled to match the new time scale. 

Now that we have a
timestep in the step function, it makes sense to have the time in there too, and
that can be updated inside the function. The data frame can have an extra
column:

```{r, eval = FALSE}
population.df <- data.frame(time = start.time,
                            susceptibles = initial.susceptibles,
                            infecteds = initial.infecteds)
```

And the `timestep_deterministic_SIS()` function can update the time along with 
the number of susceptibles and infecteds. Here, it makes the code slightly more
elegant but it might be critical if, for instance, we wanted to adjust the 
timestep as we went along, which we will need to do later. Now we’re not making 
the vector of timesteps in advance, we can use a `while()` loop to decide when 
we’re finished. For instance:

```{r, eval = FALSE}
latest.population <- population.df
while (latest.population$time < end.time) {
  latest.population <- timestep_deterministic_SIS(latest = latest.population,
                                                  transmission.rate = ecoli.transmission, 
                                                  recovery.rate = ecoli.recovery, 
                                                  timestep = this.timestep)
  population.df <- rbind(population.df, latest.population) 
}
```

`latest.population$time` is the time that we are currently at, which we extract
from the data frame after each timestep.

## Running the code

Run this code and check that it gives the same answers as in the previous 
practical. Don’t forget to check that your function is using only arguments 
passed to it by running `findGlobals()`.

## Report

Demonstrate that the code is the same in a report by running both
`step_deterministic_SIS()` and `timestep_deterministic_SIS()` with a couple of
different values of timestep (using `source()` to load in both functions from 
their respective files).
