---
title: "Reproducible Programming in R Practical 2-3"
output:
  learnr::tutorial:
    theme: lumen
runtime: shiny_prerendered
description: "Reproducible Programming in R Practical 2-3!"
---

```{r setup, include=FALSE}
library(learnr)

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```

## Overview
*In this session you will make it easier to reuse your code by placing functions in separate files, so that they can be referred to directly without having to be copied and pasted from file to file – a potentially error-strewn pastime.*

## Background

Breaking down the program into separate files is an important habit to get into. 
It makes the code less prone to error in the longer term and easier to debug 
because by creating independent units we are creating generic pieces of code 
that we can re-use over and over again in other programs. Not only is reusing 
code by copying and pasting code from one file to another inherently dangerous 
as you may accidentally copy the wrong piece of code, but if a mistake is 
identified at a later date there may be several place where the mistake has been 
copied to and you may not remember them all, causing avoidable mistakes further 
down the line.

## Tasks

The task for this practical is hopefully very simple! We are to turn the single 
program from the previous practical into two pieces of code by putting the 
`step_simple_growth()` function into a separate file, and include that file from 
the original file using the `source()` function. We have called the new file containing the function `0203-step-simple-growth.r`, so we will use the command:

```{r}
source("0203-step-simple-growth.r")
``` 

which tells our new version of the 
program, which we have called `0203-run-simple-growth.r`, that it needs to look 
in `0203-step-simple-growth.r` for functions. Your code may differ slightly 
depending on exactly what your files are called. Finish writing (and commenting) 
these two new files. Note that the function documentation should be in the 
function file now and not in the main script.

## Running the code

Now to run the code, we don’t even need to open the file 
**0203-step-simple-growth.r** – we just run **0203-run-simple-growth.r** making 
sure that we have specified **0203-step-simple-growth.r** as a source in it. Its 
contents are safely packaged away unless we want to edit them. 
**0203-run-simple- growth.r** and **0203-step-simple-growth.r** together do 
**exactly** what **0202-growth-functional.r** did. Check that your code produces 
the same results as the previous two programs.

## A clean start

Sometimes you can find that something you have done before in R can interfere 
with what you are now doing. If you’ve written all of your code correctly, then 
obviously(!) this should never happen, but you may anyway know that you can 
clear what is held in R’s memory using `rm(list=ls())` or by clicking the brush 
on the **Workspace** in **RStudio** (or <span style="color: #1697C0;">Session > Clear Workspace...</span> in the menus). All 
of the variables and functions should have now disappeared (*e.g.* if you type 
`human.annual.growth` in R it should tell you that the object is “not found”). 
Generally, I want you to produce reports though, and it turns out that when you 
generate a report, it automatically starts a new copy of R to generate it. It’s 
a good idea to regularly make reports and look at them to check they are 
formatting as you expect, so this is just another reason to do that – it’s as if 
you did the test without actually wiping everything in R.

Fundamentally, I’m not a big fan of `rm(list=ls())`, and certainly not in any of 
your scripts (so 
**don’t use it in anything you submit under any circumstances**), because it’s 
generally a sign your code isn’t written well if it can be affected by something 
before it, and in any event the workspace will be cleared when you generate the 
report. However, if things go irretrievably wrong, I would suggest you use 
<span style="color: #1697C0;">Session > Restart R</span> in the menus: it is far 
more effective than clearing just the workspace as it resets everything, while 
leaving **RStudio** itself unaffected. Remember that once you have done this you may 
have to set the working directory to the correct folder if you’re not sensibly 
using a project, and you will certainly have to start running your code again 
completely from the beginning.

## Report

As before, generate a report from the script file (hopefully you can work out 
which one this is!) and check that it produces the same results as before. Note 
again that the notebooks are always generated from a new, clean, version of R, 
so this is a very important final check that all of your code is working 
correctly – but only if you read it and check it looks right! The other check I 
strongly encourage you to do is to check that your function does not use any 
global variables by using `findGlobals()`.

Secondly, generate a report for the function file itself. If you want to, you 
can put the `findGlobals()` check into this function file, because now you are 
running this check when you generate the report for that file, so it doesn’t 
clutter up the main script itself, but this is only any good if you generate a 
report from the function file! You can make it that the report will not generate 
for that script unless you aren’t using global variables with the following code 
snippets:

```{r}
#' Now check that function doesn't have any global variables.
library(codetools)
globals <- findGlobals(step_simple_growth, merge=FALSE)$variables 
if (length(globals) != 0) {
stop("Function step_simple_growth() may not use global variable(s): ", globals)
}
```
