---
title: "Install RPiR"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
  toc: true
  toc_depth: 2
  vignette: >
    %\VignetteIndexEntry{install_RPiR}
    %\VignetteEngine{knitr::knitr}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Installation

You need to install this *Reproducible Programming in R* (RPiR) package to carry
out the exercises for this course.

## Instructions

First, make sure that the `devtools` package is installed by
calling `library(devtools)`. If that doesn't work, then install it:

```{r, eval = FALSE}
install.packages("devtools") # If need be
library(devtools)
```

Because the `RPiR` package is hosted on github, we need to use `install_github`
(from the `devtools` package):

```{r, eval = FALSE}
install_github("IBAHCM/RPiR")
```

If you see something like this:

```
These packages have more recent versions available.
Which would you like to update?
1: All                              
2: CRAN packages only               
3: None                             
4: plyr      (1.8.4 -> 1.8.5) [CRAN]
5: cli       (1.1.0 -> 2.0.0) [CRAN]
6: backports (1.1.4 -> 1.1.5) [CRAN]
```

**press 1**.. and if you're asked:

```
Do you want to install from sources the package which needs compilation?
```

say **yes**.

If this has worked, you should be able to load the `RPiR` package by typing:

```{r, eval = FALSE}
library(RPiR)
```

## Potential problems

If this didn't work, you have probably got an installation issue. The 
installation details should be printed in your RStudio Console and can help 
you troubleshoot the error. 

Quite often restarting R 
(<span style="color: #de77ae;">Session > Restart R</span>) can solve your 
problem. Just remember that once you've restarted R, you'll need to call 
`library(devtools)` again to load it back into RStudio.

### non-zero exit status

```
Warning in install.packages :
  installation of package ‘rlang’ had non-zero exit status
```

Try restarting R (<span style="color: #de77ae;">Session > Restart R</span>) and
reinstalling whatever package had non-zero exit status (zero is success, so 
non-zero is a failed install), in this case `rlang`:

```{r, eval = FALSE}
install.packages("rlang")
```

Assuming this worked, re-run `install_github("IBAHCM/RPiR")`.

### Windows

#### Rtools is required to build R packages but is not currently installed...

```
WARNING: Rtools is required to build R packages but is not currently installed. Please download and install the appropriate version of Rtools before proceeding:
```

Go here to download and install 
[Rtools](https://cran.r-project.org/bin/windows/Rtools/). 

This might also be associated with a package install error, such as:

```
namespace ‘rlang’ 0.4.0 is already loaded, but >= 0.4.1 is required
```

In which case you should reinstall whatever package had the problem,
in this case the `rlang` package:

```{r, eval = FALSE}
install.packages("rlang")
```

Assuming this worked, re-run `install_github("IBAHCM/RPiR")`.

#### cannot remove prior installation...

```
"Error in install.packages : cannot remove prior installation of package ‘rlang’ "
```

For any `cannot remove prior installation` error, try restarting R 
(<span style="color: #de77ae;">Session > Restart R</span>) and
reinstalling whatever package Windows couldn't remove, in this case `rlang`:

```{r, eval = FALSE}
install.packages("rlang")
```

Assuming this worked, re-run `install_github("IBAHCM/RPiR")`.

#### Try removing ‘C:\.../00LOCK’

```
package ‘rlang’ successfully unpacked and MD5 sums checked
Error in install.packages : ERROR: failed to lock directory ‘C:\Users\User\Documents\R\win-library\3.6’ for modifying
Try removing ‘C:\Users\User\Documents\R\win-library\3.6/00LOCK’
```

For this one, you need to find the <span style="color: #de77ae;">00LOCK</span> 
file and delete it. Then try installing whatever package  was being unpacked, 
in this case `rlang`:

```{r, eval = FALSE}
install.packages("rlang")
```

Assuming this worked, re-run `install_github("IBAHCM/RPiR")`.

### Mac / Linux

If you are on a Mac, it's very
likely you got an error because you need to have `libgit2` installed. If so,
then you probably also need to install Homebrew - follow the instructions on
https://brew.sh to do this. Once Homebrew is installed, you can
install `libgit2` using Homebrew by opening the Terminal app and then typing:

```
brew install libgit2
```

The same may happen on Linux, and you will need to use the distribution's
package manager to install either `libgit2-dev` or `libgit2-devel`, depending
on whether you are on Debian/Ubuntu or Fedora/CentOS/RHEL, respectively.
