---
title: "Install RPiR"
author: "Richard Reeve"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
  toc: true
  toc_depth: 2
  vignette: >
    %\VignetteIndexEntry{install_RPiR}
    %\VignetteEngine{knitr::knitr}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Installation

First you need to install the *Reproducible Programming in R* (RPiR) package.

## Instructions

First, make sure that the `devtools` package is installed by
calling `library(devtools)`. If that doesn't work, then install it, and then
install `RPiR` itself:

```{r, eval = FALSE}
install.packages("devtools") # If need be
library(devtools)
install_github("IBAHCM/RPiR")
```

If this has worked, you should be able to load the `RPiR` package by typing:

```{r, eval = FALSE}
library(RPiR)
```

## Potential problems

If this didn't work, you have probably got an installation issue. You can scroll
up through the pages of installation details, but if you are on a Mac, it's very
likely you got an error because you need to have `libgit2` installed. If so,
then you probably also need to install Homebrew - follow the instructions on
https://brew.sh to do this. Once Homebrew is installed, you can
install `libgit2` using Homebrew by opening the Terminal app and then typing:

```
brew install libgit2
```

The same may happen on Linux, and you will need to use the distribution's
package manager to install either `libgit2-dev` or `libgit2-devel`, depending
on whether you are on Debian/Ubuntu or Fedora/CentOS/RHEL, respectively.
