---
title: "Course preparation - RPiR"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
  toc: true
  toc_depth: 2
  vignette: >
    %\VignetteIndexEntry{Course preparation - RPiR}
    %\VignetteEngine{knitr::knitr}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

*Install the RPiR package*

You need to install the Reproducible Programming in R (RPiR) package to carry
out the exercises for this course.

## Instructions

First, make sure that the `devtools` package is installed by
calling `library(devtools)`. If that doesn't work, then install it:

```{r, eval = FALSE}
install.packages("devtools")
```

Because the `RPiR` package is hosted on github, we need to use `install_github`
(from the `devtools` package):

```{r, eval = FALSE}
library(devtools)
install_github("SBOHVM/RPiR")
```

If you see something like this:

```
These packages have more recent versions available.
Which would you like to update?
1: All                              
2: CRAN packages only               
3: None                             
4: plyr      (1.8.4 -> 1.8.5) [CRAN]
5: cli       (1.1.0 -> 2.0.0) [CRAN]
6: backports (1.1.4 -> 1.1.5) [CRAN]
```

<span style="color: #de77ae;">press 1</span>.. and if you're asked:

```
Do you want to install from sources the package which needs compilation?
```

say <span style="color: #de77ae;">no</span>.

If this has worked, you should be able to load the `RPiR` package by typing:

```{r, eval = FALSE}
library(RPiR)
```
