---
title: "R Notebook"
date: '`r format(Sys.Date(), "%B %d %Y")`'
output:
  rmarkdown::html_vignette:
  toc: true
  toc_depth: 2
  vignette: >
    %\VignetteIndexEntry{R Notebook}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::knitr}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

*Formatting R Notebook reports*

This page contains information to help you generate reports for your practical
coursework and final project. In the examples below, the contents
of your scripts are shown on the left, with the compiled report on the right.

## Headers

:::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}

::: {}
```{r}
#' # First level header
#' 
#' A nice paragraph.
#'
#' ## Second level header
#' 
#' Important information
#' 
#' ### Third level header
#'
#' More text
#'
```
:::

::: {}
# First level header {data-toc-skip=true}

A nice paragraph.

## Second level header {data-toc-skip=true}

Important information

### Third level header {data-toc-skip=true}

More text
:::

::::

## Text

You can write any text you like after the `#'` comment characters to describe
your program. If you're using *RStudio*, you can just press *Enter* every
time you get to the dotted vertical line in the window so that your R text
doesn't scroll off the page unnecessarily. You automatically get the right
comment characters so you can just carry on typing. However, there are lots
of things you can do to format the text.

For instance, you can *italicise* and **bold** text easily if you want to.
Information on the some of the basic rules for writing text (not R) are on [daringfireball][daringfireball], though most stuff is
explained below too. This format is called Markdown, and it describes
everything that happens after the `#'` symbols, though those symbols are
nothing to do with Markdown, they are just how to get *Markdown* into *R*.

You need to press enter twice to start a new paragraph, and when you're
starting a new style of text, like a list or a header you need to have a
blank comment line before it to make sure that the words don't appear in the
previous paragraph by accident.

:::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}

::: {}
```{r}
#' **first paragraph in bold**
#'
#' *second paragraph in italics*
#'
```
:::

::: {}
**first paragraph in bold**

*second paragraph in italics*
:::

::::

## Lists

:::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}

::: {}
```{r}
#' Some text followed by an empty line, 
#' then a list.
#' 
#' 1. Lists are started by numbers, 
#'    letters, or symbols (`-`, `+`, 
#'    or `*`)
#' 2. Write an equation by encapsulating
#'    an expression in `$`
#' 3. For instance:
#'    - $D = x^2$
#'    - $E = sin(q)$
#' 4. Note that you need an indent to
#'    make a sub-list
#'    a. Like this!
#'
#' More text after an empty line.
#'
```
:::

::: {}
Some text followed by an empty line, then a list.

1. Lists are started by numbers, letters, or symbols (`-`,
`+` or `*`)
2. Write an equation by encapsulating an expression in `$`
3. For instance:
   - $D = x^2$
   - $E = sin(q)$
4. Note that you need an indent to make a sub-list
   a. Like this!

More text after an empty line.
:::

::::

## Code

R comments just appear as normal, but outputs appear with `##` before them.

:::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}

::: {}
```{r, eval = FALSE}
# A useful comment
x <- 2
y <- (3 + x)^2
y
```
:::

::: {}
```{r}
# A useful comment
x <- 2
y <- (3 + x)^2
y
```
:::

::::

If you don't like that, you can remove the `##` before the output like this:

:::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}

::: {}
```{r, eval = FALSE}
#+ comment=""

x <- 2
x + 3
```
:::

::: {}
```{r, comment=""}
x <- 2
x + 3
```
:::

::::

> **Note:** If you want to remove the `##` from your entire report, you can put
`knitr::opts_chunk$set(comment = "")` at the top of your script.

:::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}

::: {}
```{r, eval = FALSE}
#' Text can also include R, so 
#' $y = (3+x)^2$ =
{{y}}
#' gives you the number calculated in R.
#'
```
:::

::: {}
Text can also include R, so $y = (3+x)^2$ =
`25`
gives you the number calculated in R.
:::

::::

## Figures

Here's a very basic figure generated using the base R `plot()` function:

:::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}

::: {}
```{r, eval = FALSE}
q <- 0:100
D <- sin(q)
plot(q, D)
```
:::

::: {}
```{r, show="hide"}
q <- 0:100
D <- sin(q)
plot(q, D)
```
:::

::::

You can specify the size of the plot and add a caption like this:

:::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}

::: {}
```{r, eval = FALSE}
#+ fig.width=2.5, fig.height=2.5, 
#+ fig.cap="Figure 2: Some sine curves"

E <- sin(q/10)
plot(q, E, type='l')
lines(q, D, col=2, lty=2)
```
:::

::: {}
```{r}
#| fig.width=2.5,
#| fig.height=2.5,
#| fig.cap="Figure 2: Some sine curves"
E <- sin(q/10)
plot(q, E, type='l')
lines(q, D, col=2, lty=2)
```
:::

::::

> **Note:** The `#+` options don't need to be split over multiple lines. We did
that here to avoid text wrapping in the code block.

> **Note also:** You can centre the figure using `#+ fig.align="center"`. We 
haven't shown that here because it doesn't work with the two-column formatting.

### `plot_populations()`

The `RPiR` package uses the `plot_simple()` and `plot_populations()` functions 
that automate some of this functionality. In addition, base `plot()` arguments
can be passed to for extra customization, see `?base::plot` for more details. 
For example, you can generate a plot like this:

```{r, fig.align="center"}
library(RPiR)

results1 <- data.frame(time = 0:100, growth = exp((0:100) / 10))
plot_populations(results1)
```

Add an additional line with a different style, width, and colour, like this:

```{r, fig.align="center"}
results1 <- data.frame(time = 0:100, growth = exp((0:100) / 10))
results2 <- data.frame(time = 0:100, death = exp(seq(10, 0, by = -0.1)))

plot_populations(results1)
plot_populations(results2, lty = 2, lwd = 3, col = "red", new.graph = FALSE)
```

Then fix the legend for the multi-line plot (making sure the order of the 
elements in the `legend`, `col`, and `lty` arguments matches the order in which 
you plot the lines!), like this :

```{r, fig.align="center"}
results1 <- data.frame(time = 0:100, growth = exp((0:100) / 10))
results2 <- data.frame(time = 0:100, death = exp(seq(10, 0, by = -0.1)))

plot_populations(results1, with.legend = FALSE)
plot_populations(results2, lty = 2, lwd = 3, col = "red", new.graph = FALSE)

legend("topright", 
       legend = c("growth", "death"), 
       lty = c(1, 2),
       lwd = c(1, 3),
       col = c("black", "red"))
```

> **Note:** You can specify the location of the legend using x-y coordinates
or strings such as `"topright"`, see `?legend` for more details.

### Potential problems

When compiling your report, you might encounter an error along the lines of:

```
Error in plot.xy(xy.coorrds(x, y), type = type, ...): plot.new has not been 
called yet Calls: <Anonymous> ... plot_populations -> <Anonymous> -> 
lines.default -> plot.xy Execution halted
```

It might be because your code looks something like this:

```{r, eval = FALSE}
results1 <- data.frame(time = 0:100, growth = exp((0:100) / 10))
results2 <- data.frame(time = 0:100, death = exp(seq(10, 0, by = -0.1)))

# Generate a figure
plot_populations(results1)

#' Some nice descriptive text for my report

# Try to plot another line on the same figure
plot_populations(results2, lty = 2, lwd = 3, col = "red", new.graph = FALSE)
```

The error occured because the first call to `plot_populations()` is separated 
from the second call to `plot_populations()` by a text block `#'`. When we
compile the report, the first and second code blocks are called in isolation, 
so the second call to `plot_populations()` isn't aware of the first plots' 
existance.

To solve this problem, do this:

```{r, eval = FALSE}
results1 <- data.frame(time = 0:100, growth = exp((0:100) / 10))
results2 <- data.frame(time = 0:100, death = exp(seq(10, 0, by = -0.1)))

plot_populations(results1)
plt <- recordPlot()

#' Some nice descriptive text for my report

replayPlot(plt)
plot_populations(results2, lty = 2, lwd = 3, col = "red", new.graph = FALSE)
```

`recordPlot()` stores the plot generated in the first code chunk and saves it 
in `plt`, afterwhich `replayPlot(plt)` recovers the plot generated in the first 
code chunk, so that
`plot_populations(results2, lty = 2, lwd = 3, col = "red", new.graph = FALSE)` 
can then be called.

## Equations

:::: {style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px;"}

::: {}
```{r, eval = FALSE}
n <- 5
x <- rnorm(n)
sum(x) / n

#' Here is the equation we just  
#' calculated written out neatly, inline 
#' with the text:  
#' $\text{mean}(x) = 1/n \sum_{i=1}^n x_i$
#' 
#' We can also write an equation on its 
#' own line:
#'
#' $$y = \frac{x_i}{2x^2 + 3}$$
#'
#' and write greek symbols and accents:
#'
#' $$\gamma = \bar{\alpha} \times \beta$$
#'

# And here's another code block
mean(x) / var(x)
```
:::

::: {}
```{r}
n <- 5
x <- rnorm(n)
sum(x) / n
```

Here is the equation we just calculated written out neatly, inline with the
text: $\text{mean}(x) = 1/n \sum_{i=1}^n x_i$

We can also write an equation on its own line:
$$y = \frac{x_i}{2x^2 + 3}$$

and write greek symbols and accents:

$$\gamma = \bar{\alpha} \times \beta$$

```{r}
# And here's another code block
mean(x) / var(x)
```
:::

::::

> **Note:** This complex syntax for equations is LaTeX. If you haven't seen it 
before you don't need to use it, but there is lots of information 
on writing formulae [here][formulae], and there are some specific
things you can put in a formula [here][format].

[daringfireball]: http://daringfireball.net/projects/markdown/syntax
[formulae]: https://en.wikibooks.org/wiki/LaTeX/Mathematics
[format]: https://en.wikipedia.org/wiki/Help:Displaying_a_formula#Formatting_using_T
