---
title: "Tips and tricks"
date: '`r format(Sys.Date(), "%B %d %Y")`'
output:
  rmarkdown::html_vignette:
  toc: true
  toc_depth: 2
  vignette: >
    %\VignetteIndexEntry{Tips and tricks}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::knitr}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## General

-   Keyboard shortcuts:

    +-------------+-----------------------------+--------------------------+
    | Description | Mac                         | Windows                  |
    +=============+=============================+==========================+
    | Run lines   | `Command + Enter`           | `Ctrl + Enter`           |
    | of code     |                             |                          |
    +-------------+-----------------------------+--------------------------+
    | (Un)comment | `Command + Shift + C`       | `Ctrl + Shift + C`       |
    | lines of    |                             |                          |
    | code        |                             |                          |
    +-------------+-----------------------------+--------------------------+
    | Insert a    | `Command + Shift + R`       | `Ctrl + Shift + R`       |
    | foldable    |                             |                          |
    | comments    |                             |                          |
    | section     |                             |                          |
    +-------------+-----------------------------+--------------------------+
    | Reindent    | `Command + I`               | `Ctrl + I`               |
    | lines of    |                             |                          |
    | code        |                             |                          |
    +-------------+-----------------------------+--------------------------+
    | Reformat    | `Command + Shift + A`       | `Ctrl + Shift + A`       |
    | lines of    |                             |                          |
    | code        |                             |                          |
    +-------------+-----------------------------+--------------------------+
    | Attempt     | `Tab`                       | `Tab`                    |
    | code        |                             |                          |
    | completion  |                             |                          |
    +-------------+-----------------------------+--------------------------+
    | Clear the   | `Control + L`               | `Ctrl + L`               |
    | console     |                             |                          |
    +-------------+-----------------------------+--------------------------+
    | Insert an   | `Alt + -`                   | `Alt + -`                |
    | assignment  |                             |                          | 
    | operator,   |                             |                          | 
    | `<-`        |                             |                          | 
    +-------------+-----------------------------+--------------------------+
    | Insert a    | `Command + Shift + M`       | `Ctrl + Shift + M`       |
    | pipe        |                             |                          |
    | operator,   |                             |                          |
    | `%>%`       |                             |                          |
    +-------------+-----------------------------+--------------------------+
    | Rename a    | `Command + Shift + M + Alt` | `Ctrl + Shift + M + Alt` |
    | variable    |                             |                          |
    | in scope    |                             |                          |
    +-------------+-----------------------------+--------------------------+

    > **Note:** If you are running R 4.1+ you can switch to the native
    > pipe operator, `|>`. To do this to to Preferences \> Code and
    > check the box "Use native pipe operator".

    > **Note also:** An easy way to remember the Rename in scope shortcut is 
    > `Pipe + Alt`.

-   To encapsulate text in quotes or brackets, highlight the text, then
    type `"` or `(`.

-   To generate code used to recreate an object:

    ```{r}
    x <- data.frame(a = 2, b = 4)
    y <- dput(x)
    y
    ```

## Writing loops

-   To generate a sequence of numbers to iterate over:

    ```{r}
    data <- c(1, 33, 100, 102)
    seq_along(data)
    ```

    alternatively:

    ```{r}
    data <- data.frame(a = 1:3, b = 4:6)
    seq_len(nrow(data))
    ```

-   To print a counter to the console:

    ```{r}
    for (i in 1:3) {
      cat("\n", i)
    }
    ```

    > **Note:** `\n` moves the "print head" to a new line, whereas `\r`
    > will move it back to the start of the same line.

## Command history

-   Keyboard shortcuts (from the console):

    +------------------------+----------------------+--------------------+
    | Description            | Mac                  | Windows            |
    +========================+======================+====================+
    | Cycle through previous | `Up arrow`           |                    |
    | commands               |                      |                    |
    +------------------------+----------------------+--------------------+
    | Show previous commands | `Command + Up arrow` |                    |
    +------------------------+----------------------+--------------------+
    | Auto-complete commands | `Control + R`        |                    |  
    | (from previous)        |                      |                    |
    +------------------------+----------------------+--------------------+
    
## Writing functions

-   To generate a function skeleton as a code snippet, type `fun` then
    hit tab and return (to select the snippet):

    ```{r}
    name <- function(variables) {

    }
    ```

    Keep hitting tab to move from `name` to `variables` to the `body` of
    the function.
    
-   Function extraction -- to turn a chunk of code into a function, highlight
    a chunk of code, then press `Control + Alt + X` on Mac (or Windows) and
    enter a suitable function name when prompted. In doing so, this:

    ```{r, eval = FALSE}
    x * y
    ```

    turns into this:

    ```{r}
    multiply <- function(x, y) {
    x * y
    }
    ```

    > **Note:** You can make your own custom code snippets in Tools > Global 
    > Options > Code and clicking Edit Snippets in the Snippets section.

## Graphics

-   To generate a sequence of evenly spaced values (useful when manually
    defining axis breaks in `ggplot2`):

    ```{r}
    x <- c(0, 0.001, 1, 8)
    pretty(x)
    ```

-   To list the names of built in colours:

    ```{r}
    x <- colors()
    head(x)
    ```

## Packages

-   Keyboard shortcuts:

    +------------------+-----------------------+------------------------+
    | Description      | Mac                   | Windows                |
    +==================+=======================+========================+
    | Document package | `Command + Shift + D` | `Ctrl + Shift + D`     |
    +------------------+-----------------------+------------------------+

## Search

-   To select the path of a file in your current directory, hit Tab
    inside of:

    ```{r, eval = FALSE}
    ""
    ```

-   To locate files relative to your project root, after installing the
    `here` package:

    ```{r}
    library(here)
    here("R", "assert_no_globals.R")
    ```

-   Keyboard shortcuts:

    +-------------------------+-----------------------+--------------------+
    | Description             | Mac                   | Windows            |
    +=========================+=======================+====================+
    | Find files / functions  | `Control + .`         | `Ctrl + .`         |
    +-------------------------+-----------------------+--------------------+
    | Search all files in a   | `Command + Shift + F` |                    |
    | directory / project     |                       |                    |
    +-------------------------+-----------------------+--------------------+

## Miscellaneous

-   Keyboard shortcuts:

    +-----------------------+-------------------------+--------------------+
    | Description           | Mac                     | Windows            |
    +=======================+=========================+====================+
    | Show command palette  | `Command + Shift + P`   | `Ctrl + Shift + P` |
    +-----------------------+-------------------------+--------------------+
    | Keyboard shortcut     | `Alt + Shift + K`       |                    |
    | reference             |                         |                    |
    +-----------------------+-------------------------+--------------------+
    | Zoom in on an RStudio | `Control + Shift + 1-4` |                    |
    | panel                 |                         |                    |
    +-----------------------+-------------------------+--------------------+
    
-   To find RMarkdown cheat sheets in RStudio, go to Help \> Cheat
    Sheets \> R Markdown Reference Guide. You can also find cheat sheets
    for `dplyr`, `ggplot2`, and `devtools` here.

-   To automatically load packages on startup / restarting R:

    1.  Create an `.Rprofile` file in your project root
    2.  Fill it with anything you want loaded on startup

    > Note that you should not use this to load in packages used in your
    > analyses, such as `dplyr` or `ggplot2`.

    A useful example might be:

    ```{r}
    library(devtools, quietly = TRUE)
    ```

    However, you can also populate it with any helper functions you
    might regularly use, or even R fortunes.

-   To generate an R fortune, after installing `fortunes`:

    ```{r, eval = FALSE}
    fortune()
    ```
