---
title: "Practical 0-1: Functions and loops"
output:
  learnr::tutorial:
    theme: lumen
    includes:
      in_header: !expr system.file("tutorials/google-font.html", package = "RPiR")
runtime: shiny_prerendered
description: >
  Students learn how to write a function and a loop.
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.reveal_solution = FALSE)
gradethis::gradethis_setup()

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```

## Overview

*In this practical you will learn a few basics of writing a `function` to wrap
 up a snippet of code you will use again, and a `for` loop that allows you to do
 the same work multiple times.*

## Introduction

If you need to use the same lines of code more than once, then you probably
need to do one of two things:

1. **If you need to do the same thing multiple times as you solve different
   problems, then you need to write a function.** Functions are a fundamental
   building block of nearly all programming languages. You use them all of the
   time, from plotting figures to building linear models, and it's not that much
   harder to write entirely new ones from scratch than it is to string a couple
   together to achieve a more specific goal than any existing function can do on
   its own.
2. **If you need to do the same thing over and over again in one place, then
   you need to write a loop.** The simplest kind of loop is a `for` loop, that
   says that you need to do a specific series of instructions an exact number
   of times.
   
It may well be that you need to do both (write functions that you then use
inside loops) - in fact you will very soon - but we'll worry about that in
a later practical.

We write both functions and loops to avoid having to copy and paste the same bit
of code over and over again into multiple different places. As well as being
deeply inelegant, this is very prone to error when you copy something slightly
wrong - you miss off a line or a letter, for instance. It is even more annoying
that if you find there is something wrong with the code you're copying, you have
to track down everywhere you copied it to and change it in all of these places.
Almost nobody ever succeeds in doing this right, and **writing functions and
loops is therefore one of the easiest ways that you can start to make your code
less likely to have mistakes in it, and therefore make the results of your work
more likely to be reproducible**.

When you are using functions and loops correctly, you should only have to fix
the code in one place, and it will automatically be fixed "everywhere", because
the code is actually only written once, it's just referenced elsewhere.

## Functions

It's very easy to write a function. Here we write a single function
(called `function_name`), that takes one input (we call the inputs to
functions *arguments*, here there is just one called `argument`). Finally,
whatever appears last in the function definition will be returned to script
that calls it, in this case, `result`.

Every function has one "line" of code after it (comments are ignored) that
defines what the function does. Because nearly all functions need multiple lines
to define what they do, we nearly always matching curly brackets - `{ ... }` -
to allow us to write multiple commands that are all treated as one line by R:

```{r intro1, eval=FALSE}
function_name <- function(argument) {
  # Function body
  ... exciting code using argument
      to do something useful, and
      producing some result ...
  result
}
```

Above, we just have dummy code, and it doesn't work. We can write a couple of
very simple functions though that really do something (not very exciting!):

```{r simple, exercise.cap = "Example 1", exercise=TRUE}
add_one <- function(number)
  number + 1

add_one(10)
```

Try running the code block by clicking on *RUN CODE*. Can you see what it does
yet? Feel free to change the `10` to make sure it works. And change what the
function does - when you do, change the name though, so it makes sense... it's
really important for your functions to have good names so it's easy to tell
if you're using them right.

The last (and indeed only) line of code in this function is what is returned by
the function, here just the number plus one. Normally we need to do something
more complex, so from now on all of our functions will use curly brackets -
`{ ... }` - to allow us to write multiple lines in the function:

```{r multiple, exercise.cap = "Example 2", exercise=TRUE}
add_one <- function(number) {
  one.more <- number + 1
  one.more
}

add_one(10)
```

Check that this does the same job. I could just have put `number + 1` inside the
curly brackets, and it would have worked the same, as this:

```{r one_line, exercise.cap = "Example 3", exercise=TRUE}
add_one <- function(number) {
  number + 1
}

add_one(10)
```

However, it's usually helpful when writing functions to name the thing that you
calculate to make it easier to keep track of what's going on. You can then
return the variable (in *Example 2*, `one.more`) that you have just calculated.
In *Example 2* you can see something very important about functions for the
first time. Try running the next example:

```{r error, exercise.cap = "Exercise 1", exercise=TRUE}
add_one <- function(number) {
  one.more <- number + 1
  one.more
}

add_one(10)
one.more
```

```{r error-solution}
add_one <- function(number) {
  one.more <- number + 1
  one.more
}

one.more <- 20
add_one(10)
one.more
```

```{r error-check}
grade_this_code()
```

You'll see that it gives an error. That is because, although you have calculated
`one.more` inside the function, it is discarded when the function ends. This is
because variables defined inside functions are *local* to the function.
For there to be a variable in your main script called one.more, you need to
define it there. Try adding a line of code between lines 5 and 6 of *Exercise 1*
that just says:

```{r, eval=FALSE}
one.more <- 20
```

Note that the *global* `one.more` variable (the one in your main script) is not
affected by calling the `add_one()` function even though it looks like it is
updated there.

## Naming of things

### Dots and underscores

You may have noticed that my functions were called
things like `add_one` (with an underscore), but my variables had names like
`one.more` (with a dot). This was deliberate. Variables and functions should be
made up of one or more words (or abbreviations) to make it easy to understand
what they do. However, it can be hard to read names just stuck together like
`addtwonumberstogether`, whereas `add_two_numbers_together` is easier.
However, there are boring reasons to do with how R works why it's a bad idea to
write R functions with dots in their names (even though many functions in R
already have dots). As a result, in these practicals, I insist that you don't
uses dots in function names when you write them.

However, dots are generally okay in variables names, so to make it easy to
distinguish normal variables from functions, I choose to put dots in variable
names (as `one.more`). You can do whatever you like.

Here's a function that takes two arguments and adds them together (yes, it
is just reinventing `+`!):

```{r rename, exercise.cap = "Exercise 2", exercise=TRUE}
add.up <- function(first, second) {
  added_together <- first + second
  added_together
}

add.up(10, 30)
```

```{r rename-solution}
add_up <- function(first, second) {
  added.together <- first + second
  added.together
}

add_up(10, 30)
```

```{r rename-check}
grade_this_code()
```

Rename the function and variables in *Exercise 2* so that it satisfies my
naming rules.

### Not reusing names

In *Exercise 1*, we showed that variables used in functions don't affect what
happens outside the function. Unfortunately, the reverse isn't true. Look at
this function:

```{r global, exercise.cap = "Exercise 3", exercise=TRUE}
missing_add <- function(first) {
  first + second
}

missing_add(10)
```

This function was intended to be like `add_up()` in *Exercise 2*, but I
accidentally forgot to add the second argument. Fortunately if you run it, it
will give an error (try it). However, what happens if you have defined
a variable called `second` elsewhere in your script? Try adding the following
at line 4:

```{r, eval=FALSE}
second <- 1
```

`missing_add(10)` should no longer give an error, and return `11`. Try changing
`second` to 2 - `missing_add(10)` should now return `12`. That's not good news -
what the function does depends on what else you may have done in your script.
When you call a function it should always do the same if you give it the same
arguments.

This gets even worse when you account for our ability to misspell things:

```{r misspelling, exercise.cap = "Exercise 3", exercise=TRUE}
misspelled_add <- function(first, second) {
  resuIt <- first + second
  result
}

result <- -5
misspelled_add(10, 20)
misspelled_add(1, 2)
```

<div id="misspelling-hint">
**Hint:** If you can't see the problem see how the word `result` is spelled in
different places.
</div>

You should find that `misspelled_add()` gives the wrong answer even though the
code looks right (if you don't look too closely). This problem can (and does!)
cause hours of agony to some students every year on this course. As a result,
we insist on two things in your submitted practicals:

1. You use variable names inside functions that you don't use outside in your
   scripts wherever possible. Sometimes there really only is one sensible name
   for a variable (maybe `timestep` when describing the increments in your
   simulation model), and you should be very careful when that happens, but
   normally there are at least two ways of describing the same idea - often one
   more general in a function (`birth.rate`) which can be used in multiple
   contexts, and then specific in your main script each time you use it
   (`human.annual.birth.rate`, for instance).
2. You write checks for your functions to make sure they aren't accidentally
   missing any variables or arguments. There are two tools for this,
   `findGlobals()` and `checkUsage()` in the `codetools` library.

Try adding the following lines to the end of *Exercise 3*:

```{r, eval=FALSE}
library(codetools)
checkUsage(misspelled_add)
```

You'll see that it reports on the problem that you have probably already
identified. It can also identify other problems, and we recommend it to you.
However, it doesn't return anything you can use in your code to allow us to
automatically reject bad functions, so there is a second solution we'll
use in our code, and we'll show you how to use it in a later exercise, but
for now just try using this line instead of `checkUsage()` above (the
`library()` call is the same):

```{r, eval=FALSE}
findGlobals(misspelled_add, merge = FALSE)
```

You'll see that it identifies the use of the global variable `result` too
(if less elegantly). We'll show you how to use this to automatically detect
problems with variable misnaming in a later exercise, but for now just remember
that you need to watch out for misspelling very carefully.

*If you're interested, the `findGlobals()` output also tells you why this
problem exists in R. You'll see that the function `misspelled_add()` needs
to use three other *global* variables - the functions `{`, `+` and `<-` - to get
the function to work. As a result of the way R works, functions have to be able
to see what is going on outside to be able to work at all. Most languages handle
this more elegantly.*

## Loops

Loops allow you to do the same thing multiple times easily. You *loop* over the
same bit of code again and again instead of having to copy and paste the code
multiple times into your script. We'll talk more about different kinds of loops
in a later practical, but for now we'll focus on the simplest - something that
tells you to do something a fixed number of times that you know in advance.
This is called a `for` loop.

A `for` loop always has the same style:

```{r, eval=FALSE}
for (one in many) {
  # Do something
    ... exciting code maybe using the variable
        one (but it doesn't have to) ...
}
```

It looks a bit like a function, but what it does is set the variable one to
each of the values in many one at a time, and then does something. Here are two
very simple examples to get an idea of what it might do:

```{r simple_loop, exercise.cap = "Example 4", exercise=TRUE}
words <- c("Hello", "Goodbye")
for (word in words) {
  print(word)
}
```

Try changing the `words` variable. You'll see that the loop runs once for each
word in the `words` vector. We can use this to do something a fixed number of
times without necessarily caring about the entries:

```{r times_loop, exercise.cap = "Example 5", exercise=TRUE}
# Make a sequence of 7 numbers from 1 to 7
times <- seq(from = 1, to = 7) # This is the same as 1:7 or seq(1, 7)

# Print "I'm in a loop" repeatedly
for (iteration in times) {
  print("I'm in a loop")
}
```

Change the length of the `times` vector. You'll see that it prints out the
text once for each element in `times` even though it never looks to see what
`iteration` is (it doesn't care).

Now look at this code:

```{r no_loop, exercise.cap = "Exercise 4", exercise=TRUE}
# A function to add one to a number
add_one <- function(number)
  number + 1

# Initialise the value
current.value <- 0

# This should be a loop!
current.value <- add_one(current.value)
current.value <- add_one(current.value)
current.value <- add_one(current.value)
current.value <- add_one(current.value)
current.value <- add_one(current.value)

# Output the updated current.value at the end to see what happened
current.value
```

```{r no_loop-check}
grade_result(pass_if(~identical(.result, 10)))
```

It just runs our `add_one()` function from earlier five times to update the
variable `current.value` to 5. Change the code so it does this ten times to
produce 10.

Now look at this next piece of code:

```{r with_loop, exercise.cap = "Exercise 5", exercise=TRUE}
# A function to add one to a number
add_one <- function(number)
  number + 1

# Initialise the value
current.value <- 0

# This now is a loop!
for (loop in 1:10) {
  current.value <- add_one(current.value)
}

# Output the updated current.value at the end to see what happened
current.value
```

```{r with_loop-check}
grade_result(pass_if(~identical(.result, 20)))
```

It does the same thing - it calls `add_one()` ten times, but it's much shorter,
which generally makes it easier to see what's going on. Now change the code
to call `add_one()` 100 times. Hopefully you can see that this is really
starting to help now!

And you can probably imagine that if it was any more complicated that just
copying and pasting one line it would make it much easier to see that it was
doing exactly the same thing each time too.

*Note also that unlike functions, variables changed in `for` loops do change
what happens outside the loop, so `current.value` is `100` at the end, and not
still `0`.*

## Conclusion

You've now learnt the basics of functions and loops. You'll be using them
throughout the rest of the practicals, so do ask for help if anything isn't
clear.
