---
title: "Practical 4-2: Add a package demo"
output: 
  learnr::tutorial:
    theme: lumen
    includes:
      in_header: !expr system.file("tutorials/google-font.html", package = "RPiR")
runtime: shiny_prerendered
description: >
  Students write a demo for their data package.
---

```{r setup, include=FALSE}
library(learnr)

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```


## Overview

*In this session you will take the biodiversity data package you have created
and write a demo that calculates a variety of diversity measures on the data.*

## Background

Now you have a package with your data in it, we can easily run analyses, simply
starting each analysis script with the R command `library(githubusernameBCI)`. 

There are an enormous number of different ways of measuring diversity, to the 
extent that diversity itself has been declared a "non-concept" in the past (by
Hurlbert, 1971), but nowadays they fit broadly within Whittaker’s framework of 
three basic types of diversity: alpha, beta and gamma diversity.

* Gamma diversity is the diversity of a metacommunity, the whole area being 
studied, whether that is an ecosystem or a single study site;
* Alpha diversity is the average diversity of the subcommunities that together
comprise the metacommunity. The metacommunity can either be broken up into
subcommunities spatially in the case of a landscape formed of several habitats,
or temporally in the case of time series data from a single site;
* Beta diversity is some measure of the differences between these 
subcommunities. This is the most variable in definition of the three types of
measure, and includes such concepts as turnover of species along a transect in
space or time, the effective number of distinct subcommunities in a 
metacommunity, the average distinctiveness of the constituent subcommunities,
and many other related concepts. Tuomisto (2010) has made a brave attempt to
make sense of these different meanings.

All of these types of diversity can be measured in several different ways. The
standard approach is often called "species diversity", where we treat all 
species as distinct and all of our measures of alpha and gamma diversity are 
counting in terms of **effective numbers of species**. However, we often need to 
consider similarity between species, and this results in a wide variety of
measures, from closely related ones such as counts of "distinct species"
(species that are taxonomically unrelated) or "distinct evolutionary histories"
(species that have no common ancestor in a phylogeny), to ones that are more
peripherally connected such as "distinct functional groups" (species that are
functionally distinct). Leinster and Cobbold (2012) have developed a framework 
for working with all of these types of similarity consistently. Because beta 
diversity measures are comparing subcommunities, they are generally unitless
(measuring fractions or percentages) or are measured in terms of numbers of
subcommunities; how similar two subcommunities are seen to be still entirely 
depends on the type of similarity though, resulting in, for instance a count of
"the effective number of phylogenetically / functionally / taxonomically
distinct subcommunities". The relationship between alpha, beta and gamma has 
been the source of many arguments – see, for example, Whittaker (1972), Jost
(2007) and Reeve et al. (2016).

Finally, I used the phrase "effective number" several times above. At one 
extreme, when we disregard abundance and just look at presence-absence data,
we measure "species richness" by simply counting the number of species present;
at the opposite extreme, we have measures such as Berger-Parker which only look
at dominance of the most abundant species. In between, we find measures such as
Simpson and Shannon, where we weight our counting of species by their abundance. 
For instance, Shannon entropy is a measure of the "surprise" or uncertainty 
expected from the next observation, which is low if there are few species or one
is dominant, but high when many are equally abundant. In all cases, the highest
diversity is achieved when all species are equally abundant (and completely 
distinct, when we are considering similarity). We therefore talk about the 
"effective number of species" as:

>the number of completely distinct, equally abundant species that would give us 
the same value of diversity as the observed community. 

This acts as a reference to allow fair comparisons to be
made. In the case of species richness, this is just the richness itself, but for
instance in the case of Shannon entropy, the *effective numbers equivalent* is 
in fact the exponential of the entropy. For diversity measures that work in 
terms of these effective numbers, the extent to which we weight by abundance is
determined by a parameter $q$, which varies from 0, where relative abundance is 
ignored and only presence-absence is considered, to $\infty$, where we 
only consider the most abundant species. This standard for dealing with 
diversities as effective numbers, while far from universal, has now been pursued 
for over 40 years and is broadly accepted – originally by Hill (1972), and
subsequently, for instance, by Jost (2007), Leinster and Cobbold (2012), and 
Reeve et al. (2016).

## Tasks

### Setting up R

First, remember as you add in `library(xxx)` calls into your demo, you'll need to
add them in as dependencies to the package using `usethis::use_package("xxx")`,
otherwise other people may not be able to run the demo. To test the demo as you
go along you may also need to install the packages in advance using
`install.packages("xxx")`.

### Create the demo files

As in the last practical series, we're going to create a **demo** inside the
package. In the <span style="color: #de77ae;">Files</span> tab, you may need to
create a new <span style="color: #de77ae;">demo</span> folder. Then create an R
script in that folder to contain your demo, call it
<span style="color: #de77ae;">diversity-of-BCI.R</span>. Note in general you
just have to make sure the name of the script has **no spaces in it**. Then
(again, as before) create a new text file also in that folder called
<span style="color: #de77ae;">00Index</span>,
with **no file extension** to contain the list of the demos in your
package. As you should know by now, each demo you create is entered into one
line of this file with the demo name followed by a one-line description. For
example, here, the demo file 
<span style="color: #de77ae;">diversity-of-BCI.R</span> would have a line like:

```
diversity-of-BCI    Showing the diversity of the BCI forest plot.
```

For more details on how to create a demo go [here](https://r-pkgs.org/misc.html#demo).

### Create the demo itself

Now you're going to write the R script that will be the demo. The script will be
measuring several kinds of alpha, beta and gamma diversity, as detailed below.
But in order to do some work with measures of taxonomic diversity, the taxonomy
will first need to match the species abundance data from the BCI dataset,
exactly. Here I’m assuming that you created the package {githubusernameBCI}, and
in it `bci_taxa` contains the taxonomy of  all BCI species and `bci_2010`
contains the 2010 BCI census.

```{r eval = FALSE}
library(githubusernameBCI)
library(dplyr)
# First filter for the correct species, either using dplyr: 
taxonomy <- bci_taxa %>%
  filter(GenusSpecies %in% rownames(bci_2010)) %>%
  unique() %>% as.data.frame()
# or using convential code:
taxonomy <- as.data.frame(unique(bci_taxa[bci_taxa$GenusSpecies %in% rownames(bci_2010),]))
# Then order the species correctly:
rownames(taxonomy) <- taxonomy$GenusSpecies 
taxonomy <- taxonomy[rownames(bci_2010),]
```

To create a metacommunity to analyse in the {rdiversity} package, you will need 
to create an object of that type using `metacommunity()`, and then you can 
calculate, for example, the gamma diversity of the whole metacommunity with the
function `meta_gamma()` and plot the output with `plot()`:

```{r eval = FALSE}
library(rdiversity)
meta <- metacommunity(bci_2010)
results <- meta_gamma(meta, qs = seq(from = 0, to = 5)) 
plot(diversity ~ q, type = "l", data = results)
```

To create a metacommunity that takes account of the similarity between species, 
you need to input the similarity as a second argument to `metacommunity()`. 
First use `tax2dist()` to create a distance matrix from a taxonomy, then use 
`dist2sim()` to create a similarity matrix from the distance matrix:

```{r eval = FALSE}
taxSim <- tax2dist(taxonomy, c(GenusSpecies=0, Genus=1, Family=2, Other=3)) %>%
  dist2sim("linear") 
metatax <- metacommunity(bci_2010, taxSim)
results4tax <- meta_gamma(metatax, qs = seq(from = 0, to = 5)) 
lines(diversity ~ q, col=2, data = results4tax)
```

Finally, to create a metacommunity to analyse in {vegan}, you need to transpose 
the axes of the abundance data matrix:

```{r eval = FALSE}
library(vegan)
# Swap rows and columns 
abundances <- t(bci_2010)
# Calculate species richness (gamma diversity)
specnumber(abundances, groups = 1) 
```

This loads the dataset and then calculates the species richness of the whole
metacommunity. **It’s important to always use the `abundances` object when using
the {vegan} package, and always use the `bci_2010` object when using
{rdiversity} –- they need the data in opposite formats!**

### Your own additions

First, read the {rdiversity} package documentation and examples. You can find
this by going to the GitHub repository (https://github.com/boydorr/rdiversity), 
and clicking on the <span style="color: #1697C0;">docs: rdiversity</span> badge, 
or by clicking the link to https://boydorr.github.io/rdiversity/.

Then write a script to calculate and plot the gamma diversity and the normalised 
alpha diversity for the 2010 BCI census, for a range of values of $q$, say 
`qs = 0:5`; individuals of the same species have a taxonomic distance of zero 
and each species is taxonomically distinct. Then repeat these calculations using
taxonomic similarity, showing that there are fewer taxonomically distinct 
species than there are species. Then do this again counting the effective number 
of genera and families. Note that this can be achieved by using `tax2dist()` 
with `c(GenusSpecies=0, Genus=0, Family=1, Other=1)` and 
`c(GenusSpecies=0, Genus=0, Family=0, Other=1)`, respectively – these distance
measurements say that two species in the same genus (or family) are zero
distance apart, just like two individuals in the same species. As a result, the
diversity measures following these instructions will simply count in effective
numbers of genera or families, respectively, rather than species.

Moving on to beta diversity, we now want to find out which quadrats in the BCI 
study site are the most distinct. We can measure the average distinctiveness of 
all of the quadrats in the plot using `raw_meta_beta()`, but you can also ask 
for the distinctiveness of every quadrat individually using `raw_sub_beta()`. 
Calculate that for a single value of $q$, say `qs=1`, and show a histogram of 
the distinctivenesses (the diversity column) of the subcommunities. Sort the 
data frame by distinctiveness and output just the highest values. Bearing in 
mind that the first two digits of the quadrat name are an x-coordinate, and the
second two are a y-coordinate, is there any pattern about these distinct areas?
If so write about it in the report.

A similar option in {vegan} is the function `betadiver()`. Look at the 
documentation in R for this function – you will see that it can calculate a very 
large range of different beta diversity measures – and calculate the beta 
diversities for the subcommunities using the `"j"` Jaccard and `"sor"` Sorensen 
similarity measures. This function calculates pairwise similarities between the 
quadrats as `dist` objects. You can convert these to a matrix and then calculate 
the average similarities to all of the quadrats by using `as.matrix()` and then 
`rowMeans()`. You should be able to show that some of the identified quadrats 
(with the lowest average similarity to the other quadrats) are the same, even
though the exact beta diversity measure, and indeed even its direction, is 
different.

Remember to add some text chunks into the report to describe the results for
when you compile a report rather than run the demo. Now reinstall the package
and make sure the demo runs successfully too. Don't forget to remove any html
files that have been generated and don't commit them to the repo.

## Checking each others packages

Now push all of your changes to GitHub and check each other's packages to make
sure the demo works.

### A note on generating a report from a demo

If you've installed a package directly using `devtools::install_github()`
you can still generate a report from the demo as follows without cloning the
repo:

```{r eval = FALSE}
rmarkdown::render(system.file("demo", "diversity-of-BCI.R", 
                              package = "mypartnersBCIpackage"),
                  output_dir = ".")
```

The report will be saved to the base folder of your current project (or whatever 
your current working directory is). Check that it looks okay by opening it in a
web browser (and delete the html file afterwards).
