---
title: "Practical 4-2"
output: 
  learnr::tutorial:
    theme: lumen
    includes:
      in_header: !expr system.file("tutorials/google-font.html", package = "RPiR")
runtime: shiny_prerendered
description: >
  Students learn how to write a package demo.
---

```{r setup, include=FALSE}
library(learnr)

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```


## Overview

*In this session you will take the biodiversity data package you created last week and write a demo that calculates a variety of diversity measures on the data.*

## Background

Now you have a package with your data in it, we can easily run analyses, simply
starting each analysis script with the R command `library(BCI)`. 

There are an enormous number of different ways of measuring diversity, to the 
extent that diversity itself has been declared a "non-concept" in the past (by
Hurlbert, 1971), but nowadays they fit broadly within Whittaker’s framework of 
three basic types of diversity: alpha, beta and gamma diversity.

* Gamma diversity is the diversity of a metacommunity, the whole area being 
studied, whether that is an ecosystem or a single study site;
* Alpha diversity is the average diversity of the subcommunities that together
comprise the metacommunity. The metacommunity can either be broken up into
subcommunities spatially in the case of a landscape formed of several habitats,
or temporally in the case of time series data from a single site;
* Beta diversity is some measure of the differences between these 
subcommunities. This is the most variable in definition of the three types of
measure, and includes such concepts as turnover of species along a transect in
space or time, the effective number of distinct subcommunities in a 
metacommunity, the average distinctiveness of the constituent subcommunities,
and many other related concepts. Tuomisto (2010) has made a brave attempt to
make sense of these different meanings.

All of these types of diversity can be measured in several different ways. The
standard approach is often called "species diversity", where we treat all 
species as distinct and all of our measures of alpha and gamma diversity are 
counting in terms of **effective numbers of species**. However, we often need to 
consider similarity between species, and this results in a wide variety of
measures, from closely related ones such as counts of "distinct species"
(species that are taxonomically unrelated) or "distinct evolutionary histories"
(species that have no common ancestor in a phylogeny), to ones that are more
peripherally connected such as "distinct functional groups" (species that are
functionally distinct). Leinster and Cobbold (2012) have developed a framework 
for working with all of these types of similarity consistently. Because beta 
diversity measures are comparing subcommunities, they are generally unitless
(measuring fractions or percentages) or are measured in terms of numbers of
subcommunities; how similar two subcommunities are seen to be still entirely 
depends on the type of similarity though, resulting in, for instance a count of
"the effective number of phylogenetically / functionally / taxonomically
distinct subcommunities". The relationship between alpha, beta and gamma has 
been the source of many arguments – see, for example, Whittaker (1972), Jost
(2007) and Reeve et al. (2016).

Finally, I used the phrase "effective number" several times above. At one 
extreme, when we disregard abundance and just look at presence-absence data,
we measure "species richness" by simply counting the number of species present;
at the opposite extreme, we have measures such as Berger-Parker which only look
at dominance of the most abundant species. In between, we find measures such as
Simpson and Shannon, where we weight our counting of species by their abundance. 
For instance, Shannon entropy is a measure of the "surprise" or uncertainty 
expected from the next observation, which is low if there are few species or one
is dominant, but high when many are equally abundant. In all cases, the highest
diversity is achieved when all species are equally abundant (and completely 
distinct, when we are considering similarity). We therefore talk about the 
"effective number of species" as:

>the number of completely distinct, equally abundant species that would give us 
the same value of diversity as the observed community. 

This acts as a reference to allow fair comparisons to be
made. In the case of species richness, this is just the richness itself, but for
instance in the case of Shannon entropy, the *effective numbers equivalent* is 
in fact the exponential of the entropy. For diversity measures that work in 
terms of these effective numbers, the extent to which we weight by abundance is
determined by a parameter $q$, which varies from 0, where relative abundance is 
ignored and only presence-absence is considered, to $\infty$, where we 
only consider the most abundant species. This standard for dealing with 
diversities as effective numbers, while far from universal, has now been pursued 
for over 40 years and is broadly accepted – originally by Hill (1972), and
subsequently, for instance, by Jost (2007), Leinster and Cobbold (2012), and 
Reeve et al. (2016).

## Setting up R

First, you need to go back to the project that contained your diversity data by
selecting it in the top right of the RStudio window. Then make sure that all of
the correct packages are installed to do the biodiversity analyses:

```{r eval = FALSE}
install.packages(c("vegan", "tibble", "dplyr", "magrittr", "devtools", "rdiversity"))
```

You will need to make sure that your package depends on all of these packages, 
so that they are automatically installed with your package. To do this you 
need to add them to the `Imports:` section of the 
<span style="color: #de77ae;">DESCRIPTION</span> file. Any package that you use 
in your demo script (*i.e.* any package you call `library(pkgname)` on) should 
go into the <span style="color: #de77ae;">DESCRIPTION</span> file.

Now we're going to create a **demo** inside the package. First go to the 
<span style="color: #de77ae;">Files</span> tab and create a <span style="color: #de77ae;">New Folder</span> called
<span style="color: #de77ae;">demo</span>. Then create an R script 
in that folder, making sure the name of the script has **no spaces in it** (this 
script will contain your demo). Then create a new text file also in that folder 
called <span style="color: #de77ae;">00Index</span>, with **no file extension**
(this file will contain a list of the demos in your package). Each demo you 
create is entered into one line of this file with the demo name followed by a 
one-line description. For example, the demo file 
<span style="color: #de77ae;">diversity-of-BCI.R</span> would have a line like:

```
diversity-of-BCI Showing the diversity of the BCI forest plot.
```

For more details on how to create a demo go here: https://r-pkgs.org/misc.html?q=demo#demo.

Now we're going to write the R script that will be the demo. Note that demos 
should never contain any commands to install packages, the above code is not 
part of the demo! The script will be measuring several kinds of alpha, 
beta and gamma diversity, as detailed below. But in order to do some work with 
measures of taxonomic diversity, the taxonomy will first need to match the 
species abundance data from the BCI dataset, exactly. Here I’m assuming that you
created the package {BCI}, and in it `bci_taxa` contains the taxonomy of 
all BCI species and `bci_2010` contains the 2010 BCI census.

```{r eval = FALSE}
library(BCI)
library(dplyr)
# First filter for the correct species, either using dplyr: 
taxonomy <- bci_taxa %>%
  filter(GenusSpecies %in% rownames(bci_2010)) %>%
  unique() %>% as.data.frame()
# or using convential code:
taxonomy <- as.data.frame(unique(bci_taxa[bci_taxa$GenusSpecies %in% rownames(bci_2010),]))
# Then order the species correctly:
rownames(taxonomy) <- taxonomy$GenusSpecies 
taxonomy <- taxonomy[rownames(bci_2010),]
```

To create a metacommunity to analyse in the {rdiversity} package, you will need 
to create an object of that type using `metacommunity()`, and then you can 
calculate, for example, the gamma diversity of the whole metacommunity with the
function `meta_gamma()` and plot the output with `plot()`:

```{r eval = FALSE}
library(rdiversity)
meta <- metacommunity(bci_2010)
results <- meta_gamma(meta, qs = seq(from = 0, to = 5)) 
plot(diversity ~ q, type = "l", data = results)
```

To create a metacommunity that takes account of the similarity between species, 
you need to input the similarity as a second argument to `metacommunity()`. 
First use `tax2dist()` to create a distance matrix from a taxonomy, then use 
`dist2sim()` to create a similarity matrix from the distance matrix:

```{r eval = FALSE}
taxSim <- tax2dist(taxonomy, c(GenusSpecies=0, Genus=1, Family=2, Other=3)) %>%
  dist2sim("linear") 
metatax <- metacommunity(bci_2010, taxSim)
results4tax <- meta_gamma(metatax, qs = seq(from = 0, to = 5)) 
lines(diversity ~ q, col=2, data = results4tax)
```

Finally, to create a metacommunity to analyse in {vegan}, you need to transpose 
the axes of the abundance data matrix:

```{r eval = FALSE}
library(vegan)
# Swap rows and columns 
abundances <- t(bci_2010)
# Calculate species richness (gamma diversity)
specnumber(abundances, groups = 1) 
```

This loads the dataset and then calculates the species richness of the whole
metacommunity. **It’s important to always use the `abundances` object when using the {vegan} package, and always use the `bci_2010` object when using {rdiversity} – they need the data in opposite formats!**

## Tasks

First, read the {rdiversity} package documentation and examples. You can find
this by going to the GitHub repository (https://github.com/boydorr/rdiversity), 
and clicking on the <span style="color: #1697C0;">docs: rdiversity</span> badge, 
or by clicking the link to https://boydorr.github.io/rdiversity/.

Then write a script to calculate and plot the gamma diversity and the normalised 
alpha diversity for the 2010 BCI census, for a range of values of $q$, say 
`qs = 0:5`; individuals of the same species have a taxonomic distance of zero 
and each species is taxonomically distinct. Then repeat these calculations using
taxonomic similarity, showing that there are fewer taxonomically distinct 
species than there are species. Then do this again counting the effective number 
of genera and families. Note that this can be achieved by using `tax2dist()` 
with `c(GenusSpecies=0, Genus=0, Family=1, Other=1)` and 
`c(GenusSpecies=0, Genus=0, Family=0, Other=1)`, respectively – these distance
measurements say that two species in the same genus (or family) are zero
distance apart, just like two individuals in the same species. As a result, the
diversity measures following these instructions will simply count in effective
numbers of genera or families, respectively, rather than species.

Moving on to beta diversity, we now want to find out which quadrats in the BCI 
study site are the most distinct. We can measure the average distinctiveness of 
all of the quadrats in the plot using `raw_meta_beta()`, but you can also ask 
for the distinctiveness of every quadrat individually using `raw_sub_beta()`. 
Calculate that for a single value of $q$, say `qs=1`, and show a histogram of 
the distinctivenesses (the diversity column) of the subcommunities. Sort the 
data frame by distinctiveness and output just the highest values. Bearing in 
mind that the first two digits of the quadrat name are an x-coordinate, and the
second two are a y-coordinate, is there any pattern about these distinct areas?
If so write about it in the report.

A similar option in {vegan} is the function `betadiver()`. Look at the 
documentation in R for this function – you will see that it can calculate a very 
large range of different beta diversity measures – and calculate the beta 
diversities for the subcommunities using the `"j"` Jaccard and `"sor"` Sorensen 
similarity measures. This function calculates pairwise similarities between the 
quadrats as `dist` objects. You can convert these to a matrix and then calculate 
the average similarities to all of the quadrats by using `as.matrix()` and then 
`rowMeans()`. You should be able to show that some of the identified quadrats 
(with the lowest average similarity to the other quadrats) are the same, even
though the exact beta diversity measure, and indeed even its direction, is 
different.

Remember if you want to add some plain text into a report, you just write it 
after a `#'` comment:

```{r}
#' We can make an observation here that Sorensen, Jaccard and {rdiversity}'s 
#' beta diversity measures identify overlapping quadrats as the least similar /
#' most distinct and it will appear as plain text in the report.
```

Generate a report from the script(s) – for each file go to the 
<span style="color: #de77ae;">File</span> menu, choose 
<span style="color: #de77ae;">Compile Report...</span>, select the 
<span style="color: #de77ae;">html</span> output format, and click 
<span style="color: #de77ae;">Compile</span>. A webpage should pop up in a 
viewer after a couple of seconds, and a new file should appear in the directory 
you are working in. You can check that this has really worked by opening the 
file in a web browser. At this point, you can improve the commenting on the 
document, so that it uses more of the features of the markdown format. Once 
you’re happy with it, recompile the report, and check it seems okay. Now
reinstall the package, call `library(BCI)` and run 
`demo("diversity-of-BCI")` or whatever you called the package and script – make 
sure that it runs as expected when you do this. Alternatively you can call 
`demo("diversity-of-BCI", package = "BCI")` without loading the library 
first.

## Reporting the results

You now have a demo file that will install with your package and show examples
of diversity calculations of the BCI data. You should commit all of these 
changes with an informative message (except the html file generated as the 
report, **you should delete this**) and push the committed changes to GitHub. 
Now open an issue on GitHub asking your partner/collaborator to test the package 
for you.

Your collaborator should now check the package to make sure that they can see
and run the demo successfully. 

As the collaborator, you should attempt to install your partner's package using 
`devtools::install_github()` using your partner's GitHub username and the name 
of their repository. You can then test the demo by using the `demo()` function. 
You can also generate a report from the demo as follows:

```{r eval = FALSE}
rmarkdown::render(system.file("demo/diversity-of-BCI.R", 
                              package = "mypartnersBCIpackage"),
                  output_dir = ".")
```

The report will be saved to the base folder of your current project (or whatever 
your current working directory is). Check that it looks okay by opening it in a
web browser (you can delete the html file afterwards). If everything looks okay, 
or if you identify a problem, you should reply to the issue saying what you 
find. Depending on whether there was a problem, you may need to go back and 
forth, or that may be everything finished...
