---
title: "Practical 2-1: SIS model"
output: 
  learnr::tutorial:
    theme: lumen
    includes:
      in_header: !expr system.file("tutorials/google-font.html", package = "RPiR")
runtime: shiny_prerendered
description: >
  Students adapt code from the previous practical series to a new application.
  They're also encouraged to use knitr::kable().
---

```{r setup, include=FALSE}
library(learnr)

knitr::opts_chunk$set(error = TRUE)
options(knitr.kable.NA = '')

set.seed(123)
```

## Overview

*In this practical we will write some code to run a simple epidemiological 
model -- the Susceptible-Infected-Susceptible or SIS model. From these results
we will report on the transmission of E. coli O157 in cattle.*

## Background

We have now built a simple population dynamics model, but the framework we are 
using is capable of handling much more. We will start with the simplest model of
infectious disease dynamics -- the SIS model -- that we covered in the
epidemiology lecture.

The SIS model requires two populations (of susceptible and infected animals) to
replace the single population we have used so far. Susceptibles, $S$, and
infecteds, $I$, are updated as follows:

$$S(t+1) = S(t) - \beta \times \frac{S(t) \times I(t)}{N} + \sigma \times I(t)$$
$$I(t+1) = I(t) + \beta \times \frac{S(t) \times I(t)}{N} - \sigma \times I(t)$$

where the total population, $N = S(t) + I(t)$, the sum of the number of
susceptibles and infecteds is constant.

As we said earlier, it is a good idea to use descriptive names in your code to
avoid confusion, so we will call the two populations `susceptibles` and
`infecteds` - these will be the column names in your data frame. The
two epidemiological parameters are $\beta$ (beta, the transmission rate) and
$\sigma$ (sigma, the recovery rate) -- these Greek letters are almost universally
used as the names of basic transmission and recovery rates in epidemiology.
However, unless you're doing a lot of epidemiology, this isn't obvious. In any
event, in our function definition we will give them descriptive names again,
calling them generically `transmission.rate` and `recovery.rate` in the function
definition, and in the script we will be more specific and call them
`ecoli.transmission` and `ecoli.recovery` since this is the  disease we are
going to be modelling (it also avoids reusing the same variable
names). The reason we're doing it this way round is that the function can
work for any disease model, but the script is specifically for *E Coli*.

## Tasks

### Technical setup

The first thing to do is to create a new GitHub repository and RStudio project
for this practical. Do this following the *how to* instructions on the [GitHub
documentation pages](https://ibahcm.github.io/RPiR/articles/pages/github.html)
that you used for the last practical, putting the repository in
the [IBAHCM](https://github.com/IBAHCM) organisation again, as before. Call this
project `githubusernameSeries02` (obviously use your real GitHub username!).
You'll use this single project for all of the exercises in this practical
series. If you are already in a coding group, then add the specific people that
need to be added to your repository on GitHub in **Settings** ->
**Manage access** -> **Add people** and just type in their GitHub
usernames (no @ sign). Give them read access. If you don't have a group yet,
ask the instructors which one you should join.

### Coding

Create a new R file, using <span style="color: #dd1c77;">File > New > R
Script</span>. You should adapt the code from Practical 1-5 (remember you now
have a project with all of this code in, and it is easy to have two RStudio
sessions so you can copy and refer to earlier code easily), creating 2 files -- 
<span style="color: #dd1c77;">0201-run-SIS.R</span>
and <span style="color: #dd1c77;">0201-step-SIS.R</span> -- but where in 1-5 
the data frame had only a `count` column, now you will need to have two --
`susceptibles` and `infecteds`. You **do not** need to have the total population
size (`N`, above) in there, because at every step it is just
`susceptibles + infecteds`. In fact, with no death in this model, it is just
constant. You could for instance start things up with:

```{r, eval = FALSE}
num.cattle <- 100
initial.infecteds <- 2
initial.susceptibles <- num.cattle - initial.infecteds
```

This allows us to easily keep the things we want to fixed, in this case the 
total population size, while allowing us to vary the things we are 
investigating. Writing the code to make this as easy as possible makes it harder 
to make mistakes, and that is always a good thing!

Once we've done that, we can create the data frame with those values -- this is
the initial state of the population when we start the simulation:

```{r, eval = FALSE}
herd.df <- data.frame(susceptibles = initial.susceptibles,
                      infecteds = initial.infecteds)
```

Then we call the new step function with the current population and the
parameters:

```{r, eval = FALSE}
next.population <- step_deterministic_SIS(latest = tail(herd.df, 1), 
                                          transmission.rate = ecoli.transmission,
                                          recovery.rate = ecoli.recovery)
```

If you are finding writing this code difficult, just ask, but remember that your
function should: 

* take in a data frame with the latest population sizes and the parameters of 
the model;
* calculate the next numbers of susceptibles and infecteds from the latest (most
recent) numbers and the parameters using the formulae above; and then 
* return a data frame with these updated values to the main script.

## Running the code

Run the code in <span style="color: #dd1c77;">0201-run-SIS.R</span>. Remember 
that the `source()` command means that it automatically loads in 
<span style="color: #dd1c77;">0201-step-SIS.R</span>, meaning that you 
don't need to load these files yourself. Don't forget to check that
your function is not using any global variables. Check that you get the same
results as you saw in the lecture. Adding the argument `col = c("green", "red")` 
to `plot.populations()` will allow you to change the colours of the (now) two
populations in the plot to match those from the lecture.

## The science

Remember from the lecture that the **basic reproduction number**, $R_0$, is 
given by 

$$R_0 = \frac{\beta}{\sigma}$$

By changing the transmission rate, $\beta$, try running simulations for 
different values of $R_0$. What happens when $R_0 > 1$ and when $R_0 < 1$? 
Run the simulations for different values of the transmission rate (as suggested 
in the table below) with the default value of the recovery rate, $\sigma$, set 
to $\frac{1}{3}$. Calculate and record the corresponding value of $R_0$ using 
the formula, which we can rewrite as:

```{r, eval = FALSE}
R0 <- ecoli.transmission / ecoli.recovery
```

Either using your plots or by asking R to 
print out the data frame `population` work out the proportion of the population
susceptible at equilibrium and compare it with the value of $\frac{1}{R_0}$.

```{r, echo = FALSE}
tmp <- vapply(c(1/2, 2/3, 1, 4/3), 
              function(x) as.character(MASS::fractions(x)),
              character(1))
table <- data.frame(tmp, NA, NA, NA)
col <- c("Transmission <br> rate, $\\beta$", 
         "Basic reproduction number, $R_0$",
         "Proportion susceptible <br> at equilibrium, $\\frac{S}{N}$",
         "Inverse $R_0$, <br> $\\frac{1}{R_0}$")
col <- gsub("&lt;", "<", col)
col <- gsub("&gt;", ">", col)
colnames(table) <- col
knitr::kable(table, escape = FALSE, align = c(rep('c', times = 4)))
```

Remember that throughout this series of exercises, the total population size,
$N$ is fixed, but can be easily calculated from the different parts of the
population so you never need to pass it as an argument or store it in the
data frame.

## Report
Adapt your script to generate a report which shows you running the model with at 
least a couple of different transmission rates and calculating different values
for $R_0$ from the results with some explanation of what's going on. You 
should also submit a table in Word or Excel with the above values filled in -- 
just save it in the project folder. In particular for *E. coli O157* infections 
in cattle, for timesteps in weeks, the recovery rate is about $\frac{1}{3}$ and 
the transmission rate is 1.

Optionally, try putting the table into the report (possibly using the
`kable()` function in the `knitr` package). However, the table itself 
(in any form) is sufficient for assessment.

## GitHub

In this exercise, we want you to get a couple of other people to check your code
and make sure it works for them, and we want you to check other people's code
too.

### Getting help with your code

Once you're happy with your code, commit your changes using the Git pane in
RStudio (just commit the R files - not, for instance, any
html files that were created by generating reports, though you may decide to
commit a spreadsheet with the table in as well). Don't forget then to push
the changes to GitHub and check on the website that it contains your new code.
Notify the partners we have allocated you that you have something for them to
check. If you're not sitting next to them, then you can create an issue in your
repository asking for their review and/or talk to them on Zoom - you can contact
them by tagging them in the message with `@theirusername`, and they should
receive an email and a [notification on the website](https://github.com/notifications).
They should then create an issue, telling you if they had any problems
running your version of the practical.

### Checking other people's code

Likewise, when you hear from someone else that a repo is available to check,
then open it using <span style="color: #dd1c77;">File > New Project...</span>,
then *Version Control*, *Git*, and then put in the URL from GitHub and tick
`Open in New Session`. In the new RStudio session that is opened, you can then
run their code and make sure it works. If it does not, just respond in the Issue
that they opened on GitHub explaining what went wrong (though it might be easier
to do this in person on Zoom if they are there). If it works, then you also need
to respond saying so. If you have already checked their repo once already, then
see the instructions in 2-2 for how to update your copy of it.

**Interacting like this through GitHub to help each other will count as your
engagement marks for the course.**
