---
title: "Practical 2-4: SIS model with timesteps"
output: 
  learnr::tutorial:
    theme: lumen
    includes:
      in_header: !expr system.file("tutorials/google-font.html", package = "RPiR")
runtime: shiny_prerendered
description: >
  Students improve their timestep function. They also learn how to use while() 
  loops.
---

```{r setup, include=FALSE}
library(learnr)

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```

## Overview

*In this practical we separate out the model from the simulation functionality more cleanly. In the process we make the previous set of code safer to use (and more elegant!).*

## Background

When we are programming we often make quick additions to code, as we did in the
last practical but it is good practice, if we want to keep them, to make sure
they are included in the code in a way that makes it clear what’s going on. We
had to know how the SIS model works to know that we could get away with 
multiplying the parameters with `timestep.` This use of a low-level detail of
how the model runs might be risky – we may later change the model and 
not realise this hack no longer works.

## Tasks

In the previous practical we multiplied the transmission and recovery rates up
(or down) to reflect changes in `timestep.` However, `transmission.rate` and
`recovery.rate` were not really changing, and we want the code to reflect that
rather than abusing it to do something it wasn't intended to do. Instead add a
new argument, `timestep`, into `step_deterministic_SIS()` – we could call it
`timestep_deterministic_SIS()`. Rather than submit the product 
`transmission.rate * timestep` as an argument to the simulation function, we can 
now separately pass the real rate parameters and the `timestep` separately.
This **rescaling** is now done within the function
`timestep_deterministic_SIS()` in
<span style="color: #de77ae;">0204-deterministic-SIS.R</span>. We might, for
instance, create two new variables, perhaps called `effective.transmission.rate` 
and `effective.recovery.rate` (or something shorter!) inside that function.
These new variables are just our original rates rescaled to match the new time
scale. 

Now that we have a timestep in the step function, it makes sense to have the
time in there too, and that can be updated inside the function. Indeed, the data
frame itself that is passed back and forth can have an extra column. This
should start in the new report script that you are writing (perhaps
<span style="color: #de77ae;">0204-run-SIS.R</span>), where it will be
initialised with all three columns (`time`, `susceptibles` and `infecteds`):

```{r, eval = FALSE}
population.df <- data.frame(time = start.time,
                            susceptibles = initial.susceptibles,
                            infecteds = initial.infecteds)
```

And the `timestep_deterministic_SIS()` function can update the time along with 
the number of susceptibles and infecteds. Here, it makes the code slightly more
elegant but it might be critical if, for instance, we wanted to adjust the 
timestep as we went along, which we will need to do later. Now we’re not making 
the vector of timesteps in advance, we can use a `while()` loop to decide when 
we’re finished. For instance:

```{r, eval = FALSE}
latest.population <- population.df
while (latest.population$time < end.time) {
  latest.population <- timestep_deterministic_SIS(latest = latest.population,
                                                  transmission.rate = ecoli.transmission, 
                                                  recovery.rate = ecoli.recovery, 
                                                  timestep = this.timestep)
  population.df <- rbind(population.df, latest.population) 
}
```

`latest.population$time` is the time that we are currently at, which we extract
from the data frame after each timestep.

## Running the code

Run this code and check that it gives the same answers as in the previous 
practical. Don’t forget to check that your function is using only arguments 
passed to it by running `findGlobals()`.

## Report

Demonstrate that the code is the same in a report by running both
`step_deterministic_SIS()` and `timestep_deterministic_SIS()` with a couple of
different values of timestep (using `source()` to load in both functions from 
their respective files).

## GitHub

As with the other exercises, we want you to get a couple of other people to
check your code and make sure it works for them, and we want you to check other
people's code too.

### Getting help with your code

Once you're happy with your code, commit your changes using the Git pane in
RStudio (it's best just to commit the R files for now - not, for instance, any
html files that were created by generating reports). Don't forget then to push
the changes to GitHub and check on the website that it contains your new code.
Notify the partners we have allocated you that you have something for them to
check. You can do this on Zoom, but please also create an issue in your
repository asking for their review - you can contact them by tagging them in
the message with `@theirusername`, and they should receive an email and a
[notification on the website](https://github.com/notifications). They can then
respond to this issue, telling you whether they had any problems running your 
version of the practical.

### Checking other people's code

Likewise, when you hear from someone else that a repo is available to check,
then if it's a project you have already downloaded, you should open that
project in RStudio, go to the Git pane, and click on the 
<span style="color: #de77ae;">Pull</span> button to
get the updates they have pushed to GitHub (there should be a subtle message
saying that updates are available near the button). If you're checking a new
person's work, or a new project, refer to the instructions in Practical 2-2.
Once you've checked that the new code has downloaded, you can then run their
code and make sure it works. If it does not, just respond in the Issue that they
opened on GitHub explaining what went wrong (it might help to also do
this in person on Zoom if they are there). If it works, then you also need
to respond saying so.

Interacting like this through GitHub to help each other will count as most of
your engagement marks for the course.
