---
title: "Practical 3-4: Stochastic SIS model"
output: 
  learnr::tutorial:
    theme: lumen
    includes:
      in_header: !expr system.file("tutorials/google-font.html", package = "RPiR")
runtime: shiny_prerendered
description: >
  Students run a stochastic version of the SIS model.
---

```{r setup, include=FALSE}
library(learnr)

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```

## Overview

*In this practical we develop and run our first stochastic version of the SIS model.*

## Background

We saw in the previous practicals the impact of including randomness into the
exponential model with births and deaths. Having developed the programming 
framework, it is now time to investigate the impact of randomness in the 
simplest epidemiological model -- the SIS model.

## Tasks

In this practical we will write a stochastic version of the SIS model. You have 
been provided with `run_simulation()` in the `RPiR` package, which allows some 
extra functionality over `run_simple()` to run simulations using the older
deterministic models. Pay attention in particular to writing 
the function, where the most important changes need to be made.

### 1. Write a function

First, you need to write a new function, `timestep_stochastic_SIS()`, adapted 
from `timestep_stochastic_birth_death()` to handle the SIS model rather than 
exponential growth. 

The function should be defined like this:

```{r, eval = FALSE}
timestep_stochastic_SIS <- function(latest, 
                                    transmission.rate, 
                                    recovery.rate, 
                                    timestep) {
  # Some code
  
}
```

It proceeds much as before with the deterministic SIS model, calculating the 
effective transmission rate (beta) and recovery rate (sigma) at this timestep. 
Then as with the stochastic birth death model, you need to work out whether the 
individual rate of infection or recovery per timestep is greater than 1, in 
which case it cannot be used as a probability. In this case, the actual 
individual infection rate is `effective.transmission.rate` $\times \frac{I}{N}$ 
and the individual recovery rate is `effective.recovery.rate`.

Now for the key step... you need to sample from the binomial distribution (as 
described in the lecture) to determine the number of new infecteds and 
susceptibles using `rbinom()` with the probability of "success" set to the
individual rates detailed above. Then update the populations, and if there are
no infecteds left, then the epidemic is over, so set `is.finished` to `TRUE.` 

```{r, eval = FALSE}
is.finished <- next.infecteds == 0
```

Then append the new data to the data frame as usual, and return a list with the
updated population data frame and a variable that tells us whether we are 
finished or not.

```{r, eval = FALSE}
list(updated.pop = next.population, end.experiment = is.finished)
```

### 2. Write a demo

Now write a new demo called 
<span style="color: #dd1c77;">d0304_run_SIS</span>, to demonstrate the SIS 
model. It might be useful to adapt the code you wrote in 
<span style="color: #dd1c77;">d0303_run_birth_death</span>.

### 3. Check your function

What happens when you do this when the time step is 1? You should get an error 
message saying the timestep is too large. You may still get a plot though! 
This is because R is plotting the previous value of populations held in its 
memory. This is why it's important to be aware of your workspace and clear your 
previous history when you're working on a new project (don't use 
`rm(list=ls())` though!).

> Remember that generating a report starts a new copy of R and so is a much 
better check of whether the script can run on its own.

Clear what is held in R's workspace in RStudio. Now rerun with $R_0 = 15$ and 
confirm that you don't get an output when the error message shows. Remember
that clearing your workspace and trying again is often a very useful trick when
you get unexpected results.

## Report

### 1. Compare stochastic and deterministic models

When you are happy with the structure of the code, run it a few times with a
transmission rate of 0.3, a recovery rate of 0.1, and 2 initial infecteds from 
a population size of 100. Demonstrate the variability in the results of the 
stochastic model.

You should also overlay the deterministic results from Practical 2-4. 
Noting that `run_simulation()` will work with this as well since it 
automatically detects that it just returns a dataframe, and behaves 
appropriately:

```{r, eval = FALSE}
final.populations <- run_simulation(timestep_deterministic_SIS, 
                                    initial.populations,
                                    end.time,
                                    transmission.rate = ecoli.trans,
                                    recovery.rate = ecoli.recov, 
                                    timestep = working.timestep)
```

### 2. Vary the rate of transmission and the initial number of infecteds

Try varying the transmission rate (beta) -- this will change $R_0$ -- and the 
initial number of infecteds and see how the likelihood of early extinction 
changes. 

### 3. Vary the total population size 

Try changing the total population size and determine the impact on the observed 
variability.

## Check it works

As with previous exercises, you need to check that everything works correctly --
that the package installs, and the demos and help files work and you can
generate reports from the demos -- and then we want you to get a couple of other
people in your breakout room to check your code and make sure it works for them,
and we want you to check other people's code too. We describe how to do this for
packages in GitHub in Practical 3-1 (also under *Check it works*) if you're
uncertain.

*Remember, interacting like this through GitHub to help each other will count as
most of your engagement marks for the course.*
