---
title: "Practical 1-6: Writing a report in RStudio"
output: 
  learnr::tutorial:
    theme: lumen
    includes:
      in_header: !expr system.file("tutorials/google-font.html", package = "RPiR")
runtime: shiny_prerendered
description: >
  Students learn to write a report in RStudio, combining everything they have 
  learned so far. They also learn how to empty the Global Environment and use 
  stop() to catch errors.
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.reveal_solution = FALSE)
gradethis::gradethis_setup()
 
knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```

## Overview

*In this session we will take the code for the models with just growth 
(Practical 1-4) and with births and deaths (Practical 1-5) and make a clearer 
report comparing the two.*

---

Note that from here on out you will be working in RStudio. If you don't still 
have it open, you should be able to open the new project that you created in 
Practical 1-1 using <span style="color: #dd1c77;">File > Open Project in New 
Session...</span> in RStudio or follow the instructions in 1-1 to create 
another new project if you want to do that.

To access course materials, you have two options:

1. In RStudio, click <span style="color: #dd1c77;">View > Panes > 
   Pane Layout... > then in the Files, Plots, Packages, Help, Tutorial, 
   Viewer</span> section, scroll down and check the box next to Tutorial. This 
   will give you a tutorial tab from which to access materials.
2. Open two versions of RStudio. One to run the `learnr` course (this will 
   likely have a big red stop button preventing you from being able to run 
   anything in the Console) and one to do your work in. 

## Background

In this practical, we are going to demonstrate why it's worth dividing up the
code into separate functions, and indeed separate files. Previously we wrote a
simple growth model, and a model of birth and death. We're now going to use the
two-step functions we created and run both models in a new script that compares 
them. For this final part of Practical 1, we're going to generate a report in 
RStudio using everything we've learned so far.

## A clean start

Sometimes you can find that something you have done before in R can interfere 
with what you are now doing -- though if you've written all of your code 
correctly, then obviously(!) this should never happen. It is useful then
to know that you can clear what is held in R's memory using `rm(list=ls())`. 
This will cause all of the variables and functions in your 
**Global Environment** to disappear (*e.g.* if you type `human.annual.growth` 
in the R console it should tell you that the object is "not found"). As a 
fundamental rule however, you should **never hardcode** `rm(list=ls())` 
**into any of your scripts**, and in fact you should just not use it.
It's generally a sign your code isn't written well if it can be affected by
something before it. Instead, you can clear your workspace by clicking the brush
in the <span style="color: #dd1c77;">Environment</span> tab in RStudio (or 
<span style="color: #dd1c77;">Session > Clear Workspace...</span> in the menus). 

However, generally it's easier to Restart R. This is much safer and more final.
You can do it in RStudio via <span style="color: #dd1c77;">Session > Restart R</span>
in the menus: it is far 
more effective than clearing just the workspace as it resets everything, while 
leaving RStudio itself unaffected. Remember that once you have done this you may 
have to set the working directory to the correct folder, and you will certainly 
have to start running your code again completely from the beginning.

Note that, when you generate a report from the script file, they are always 
generated from a new, clean, version of R. So this is a very important final 
check that all of your code is working correctly -- it's as if 
you did restart R without actually wiping everything in your current R console -
but only if you read it and check it looks right!

The other check we strongly encourage you to do is to check that your function 
does not use any global variables by using `findGlobals()`.

You should also generate a report for any function files you create. If you want 
to, you can put the `findGlobals()` check into this function file (which will
run the check when you generate the report for that file) so it doesn't 
clutter up the main script itself. But this is only useful if you generate a 
report from the function file! To solve this, you can make it so the report will 
fail to generate for that script when global variables are used with the 
following code snippets:

```{r, eval = FALSE}
# Now check that function doesn't have any global variables.
library(codetools)
globals <- findGlobals(step_simple_growth, merge = FALSE)$variables 
if (length(globals) != 0) {
  stop("Function step_simple_growth() may not use global variable(s): ", globals)
}
```

## Tasks

Remember again that our first model of simple population growth model looks like
this:

$$N(t+1) = N(t) + \lambda \times N(t)$$

The birth and death model is slightly more complex:

$$ N(t+1) = N(t) + b \times N(t) - d \times N(t)$$

Open RStudio and create a new script called 
<span style="color: #dd1c77;">0106-final-report.R</span>.

Remember that <span style="color: #dd1c77;">0104-step-growth.R</span> has the
`step_deterministic_growth()` function and 
<span style="color: #dd1c77;">0105-step-birth-death.R</span> has the 
`step_deterministic_birth_death()` function, so we need to `source()` both of
these files at the top of our script.

We want to compare the models, so write some code that first runs a simulation 
with one model and plots it, and then runs a simulation with the other model 
and plots that as well. Remember that you can plot a second (or third, etc.) 
result on top of a first with commands like:

```{r, eval = FALSE}
plot_populations(population.a, new.graph = TRUE, col = "black")
plot_populations(population.b, new.graph = FALSE, col = "red")
```

## Report

For this report, we want something substantially clearer and better laid out, 
showing something that demonstrates the relationship between the growth and the
birth-death models, noting for instance that the crude global birth and death 
rates are around 23 per 1000 population and 8 per 1000 respectively. You can see
an example of most of the possible formatting options in 
<span style="color: #dd1c77;">0200-example-notebook.R</span>. Use an 
understanding of what you want to describe --
and formatting from the example file or elsewhere -- to break up your script so 
that your code and the results and figures are interspersed with a reasonable
explanation of what is going on.

### 1. Vary parameters

You should try running:

1. The birth-death model with a single given birth and death rate
2. The simple growth model with the birth rate as the growth rate
3. The simple growth model with minus the death rate as the (negative) growth
rate
4. The simple growth model with the difference (between the rates in 1) as the 
growth rate

When you try to run your code you should find that the graphs are very likely on
different scales. You can set the y-axis limits of the second graph to the same 
as the first by reading off what they are on the first graph, and then plotting
the second as follows:

```{r, eval = FALSE}
plot_populations(population.df, ylim = c(0, 20000000000))
```

Once you've done this it should be easier to compare the two (by eye at least).
