---
title: "Reproducible Programming in R Practical 1-6"
output: 
  learnr::tutorial:
    theme: lumen
runtime: shiny_prerendered
description: >
  Students learn to write a report in RStudio, combining everything they have 
  learned so far. They also learn how to empty the Global Environment and use 
  stop() to catch errors.
---

```{r setup, include=FALSE}
library(learnr)
gradethis::gradethis_setup()
 
knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```

## Overview

*In this session we will take the code for the models with just growth (1-4)* 
*and with births and deaths (1-5) and make a clearer report comparing the two.*

## Background

We are now going to demonstrate in a small way why it’s worth dividing up the
code into separate functions, and indeed separate files. Previously we wrote a
simple growth model, and a model of birth and death. We’re now going to use the
two-step functions we created and run both models in a new script that compares 
them. For this final part of practical 1, we're going to generate a report in 
RStudio using everything we've learned so far.

## Tasks

Remember again that our first model of simple population growth model looks like
this:

$$N(t+1) = N(t) + \lambda\; N(t)$$

The birth and death model is slightly more complex:

$$ N(t+1) = N(t)+ b\; N(t) – d\; N(t)$$

Remember that <span style="color: #de77ae;">0104-step-growth.r</span> has the
`step_deterministic_growth()` function and 
<span style="color: #de77ae;">0105-step-birth-death.r</span> has the 
`step_deterministic_birth_death()` function, so we need to `source()` both of
these files (as well as 
<span style="color: #de77ae;">0104-plot-helper.r</span>).

Now open RStudio. For this practical, we're going to generate a report called 
<span style="color: #de77ae;">0106-final-report.r</span>. We want to compare the
models, so write some code that first runs a simulation with one model and plots 
it, and then with the other model and plots that as well. Remember that you can 
plot a second (or third, etc.) result on top of a first with commands like:

```{r, eval = FALSE}
plot_populations(population.b, new.graph = FALSE, col = "red")
```

You should try running:

1. The birth-death model with a single given birth and death rate
2. The simple growth model with the birth rate as the growth rate
3. The simple growth model with minus the death rate as the (negative) growth
rate
4. The simple growth model with the difference (between the rates in 1) as the 
growth rate

## Running the code

Try running your code and you should find that the graphs are very likely on
different scales. You can set the y-axis limits of the second graph to the same 
as the first by reading off what they are on the first graph, and then plotting
the second as follows:

```{r, eval = FALSE}
plot_populations(population.df, ylim = c(0, 20000000000))
```

Once you’ve done this it should be easier to compare the two (by eye at least).

## Report

For this report, we want something substantially clearer and better laid out, 
showing something that demonstrates the relationship between the growth and the
birth-death models, noting for instance that the crude global birth and death 
rates are around 23 per 1000 population and 8 per 1000 respectively. You can see
an example of most of the possible formatting options in 
<span style="color: #de77ae;">0100-example-notebook.r</span>. Use an 
understanding of what you want to describe –
and formatting from the example file or elsewhere – to break up your script so 
that your code and the results and figures are interspersed with a reasonable
explanation of what is going on.
