---
title: "Practical A-3: Naming of things"
output:
  learnr::tutorial:
    theme: lumen
    includes:
      in_header: !expr system.file("tutorials/google-font.html", package = "RPiR")
runtime: shiny_prerendered
description: >
  Basics of variable and function naming and global variables.
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.reveal_solution = FALSE)
gradethis::gradethis_setup()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ">",
  error = TRUE
)
set.seed(123)
```

## Naming of things

### Dots and underscores

You may have noticed that my functions in A-2 were called
things like `add_one` (with an underscore), but my variables had names like
`one.more` (with a dot). This was deliberate. Variables and functions should be
made up of one or more words (or abbreviations) to make it easy to understand
what they do. However, it can be hard to read names just stuck together like
`addtwonumberstogether`, whereas `add_two_numbers_together` is easier.
However, there are boring reasons to do with how R works why it's a bad idea to
write R functions with dots in their names (even though many functions in R
already have dots). As a result, in these practicals, I insist that you don't
uses dots in function names when you write them.

However, dots are generally okay in variables names, so to make it easy to
distinguish normal variables from functions, I choose to put dots in variable
names (as `one.more`). You can do whatever you like.

Here's a function that takes two arguments and adds them together (yes, it
is just reinventing `+`!):

```{r rename, exercise.cap = "Exercise 2", exercise=TRUE}
add.up <- function(first, second) {
  added_together <- first + second
  added_together
}

add.up(10, 30)
```

```{r rename-solution}
add_up <- function(first, second) {
  added.together <- first + second
  added.together
}

add_up(10, 30)
```

```{r rename-check}
grade_this_code()
```

Rename the function and variables in *Exercise 2* so that it satisfies my
naming rules.

*There is much more on how to style your code to make it more consistent
[here](http://adv-r.had.co.nz/Style.html). Hadley recommends never using dots,
so using underscores to separate words in functions and variables. If you want
to do that it's fine. Don't use dots in function names though. He has lots of
other suggestions about how to write clear code - do have a look at it and try
to follow it if you have time.*

### Not reusing names

In *Exercise 1*, we showed that variables used in functions don't affect what
happens outside the function. Unfortunately, the reverse isn't true. Look at
this function:

```{r global, exercise.cap = "Exercise 3", exercise=TRUE}
missing_add <- function(first) {
  first + second
}

missing_add(10)
```

This function was intended to be like `add_up()` in *Exercise 2*, but I
accidentally forgot to add the second argument. Fortunately if you run it, it
will give an error (try it). However, what happens if you have defined
a variable called `second` elsewhere in your script? Try adding the following
at line 4:

```{r, eval=FALSE}
second <- 1
```

`missing_add(10)` should no longer give an error, and return `11`. Try changing
`second` to 2 - `missing_add(10)` should now return `12`. That's not good news -
what the function does depends on what else you may have done in your script.
When you call a function it should always do the same if you give it the same
arguments.

This gets even worse when you account for our ability to misspell things:

```{r misspelling, exercise.cap = "Exercise 4", exercise=TRUE}
misspelled_add <- function(first, second) {
  resuIt <- first + second
  result
}

result <- -5
misspelled_add(10, 20)
misspelled_add(1, 2)
```

<div id="misspelling-hint">
**Hint:** If you can't see the problem see how the word `result` is spelled in
different places.
</div>

You should find that `misspelled_add()` gives the wrong answer even though the
code looks right (if you don't look too closely). This problem can (and does!)
cause hours of agony to some students every year on this course. As a result,
we insist on two things in your submitted practicals:

1. You use different variable names in your functions than in your scripts,
   wherever possible. Sometimes there really only is one sensible name
   for a variable (maybe `timestep` when describing the increments in your
   simulation model), and you should be very careful when that happens, but
   normally there are at least two ways of describing the same idea - often one
   more general in a function (`birth.rate`) which can be used in multiple
   contexts, and then specific in your main script each time you use it
   (`human.annual.birth.rate`, for instance).
2. You write checks for your functions to make sure they aren't accidentally
   missing any variables or arguments. There are two tools for this,
   `findGlobals()` and `checkUsage()` in the `codetools` library.

Try adding the following lines to the end of *Exercise 4*:

```{r, eval=FALSE}
library(codetools)
checkUsage(misspelled_add)
```

You'll see that it reports on the problem that you have probably already
identified. It can also identify other problems, and we recommend it to you.
However, it doesn't return anything you can use in your code to allow us to
automatically reject bad functions, and we'll show you how to do this in a
later practical. For now just try using this line instead of `checkUsage()`
above (the `library()` call is the same):

```{r, eval=FALSE}
findGlobals(misspelled_add, merge = FALSE)
```

You'll see that it identifies the use of the global variable `result` too
(if less elegantly). We'll show you how to use this to automatically detect
problems with variable misnaming in a later exercise, but for now just remember
that you need to watch out for misspelling very carefully.

*If you're interested, the `findGlobals()` output also tells you why this
problem exists in R. You'll see that the function `misspelled_add()` needs
to use three other *global* variables - the functions `{`, `+` and `<-` - to get
the function to work. As a result of the way R works, functions have to be able
to see what is going on outside to be able to work at all. Most languages handle
this more elegantly.*

Functions are covered in R4DS [here][functions-R4DS] and R Coder covers
them in some depth [here][functions-rcoder]. Advanced R covers functions in
enormous detail starting [here][functions-advanced-r2] with an introduction to
functional programming, and continuing for several chapters.

[functions-R4DS]: https://r4ds.had.co.nz/functions.html
[functions-advanced-r2]: https://adv-r.hadley.nz/fp.html
[functions-rcoder]: https://r-coder.com/function-r
