---
title: "Practical A-2: Loops"
output:
  learnr::tutorial:
    theme: lumen
    includes:
      in_header: !expr system.file("tutorials/google-font.html", package = "RPiR")
runtime: shiny_prerendered
description: >
  Basics of repeating instructions multiple times - for and while loops.
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.reveal_solution = FALSE)
gradethis::gradethis_setup()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ">",
  error = TRUE
)
set.seed(123)
```

## Loops 

Loops allow you to repeat the same thing again and again. If you are doing
something more than two or three times, you should instead use a loop to
avoid having to copy and paste the same code over and over again.
Since duplicating code risks not only mistyping or not copying correctly, but
also creating confusion by filling your R scripts with nearly identical 
chunks of code. There are fundamentally two kinds of loop:

- one where you know exactly how many times the loop will run, for instance 
  because you're running a chunk of code once for each element of a vector or
  row or column of a data frame; and
- one where you do not, because some part of your calculation in each step of
  the loop will determine whether to continue or to stop.

### `for` loops

`for` loops allow us to repeat something **for** a fixed number of times (*e.g.*
once per element of a vector, once per row of a data frame, *etc.*). 

A `for` loop always has the same style:

```{r, eval=FALSE}
for (one in many) {
  # Do something
  
  # ... exciting code maybe using the variable
  #     one (but it doesn't have to) ...
}
```

It looks a bit like a function, but what it does is set the variable `one` to
each of the values in `many` one at a time, and then does something. Here are two
very simple examples to get an idea of what it might do:

```{r forloop, exercise.cap = "Exercise 1", exercise = TRUE}
greetings <- c("Hello", "Goodbye")
for (word in greetings) {
  print(word)
}
```

Try changing the `greetings` variable. You'll see that the loop runs once for each
word in the `greetings` vector. We can use this to do something a fixed number of
times without necessarily caring about the entries:

```{r times_loop, exercise.cap = "Exercise 2", exercise=TRUE}
# Make a sequence of 7 numbers from 1 to 7
times <- seq(from = 1, to = 7) # This is the same as 1:7 or seq(1, 7)

# Print "I'm in a loop" repeatedly
for (iteration in times) {
  print("I'm in a loop")
}
```

Change the length of the `times` vector. You'll see that it prints out the
text once for each element in `times` even though it never looks to see what
`iteration` is (it doesn't care).

Now look at this code:

```{r no_loop, exercise.cap = "Exercise 3", exercise=TRUE}
# A function to add one to a number
add_one <- function(number)
  number + 1

# Initialise the value
current.value <- 0

current.value <- add_one(current.value)
current.value <- add_one(current.value)
current.value <- add_one(current.value)
current.value <- add_one(current.value)
current.value <- add_one(current.value)

# Output the updated current.value at the end to see what happened
current.value
```

```{r no_loop-check}
grade_result(pass_if(~identical(.result, 10)))
```

It just runs our `add_one()` function from earlier five times to update the
variable `current.value` to 5. Change the code so it does this ten times to
produce 10.

Now look at this next piece of code:

```{r with_loop, exercise.cap = "Exercise 4", exercise=TRUE}
# A function to add one to a number
add_one <- function(number) {
  number + 1
}

# Initialise the value
current.value <- 0

# This now is a loop!
for (loop in 1:10) {
  current.value <- add_one(current.value)
}

# Output the updated current.value at the end to see what happened
current.value
```

```{r with_loop-check}
grade_result(pass_if(~identical(.result, 100)))
```

It does the same thing - it calls `add_one()` ten times, but it's much shorter,
which generally makes it easier to see what's going on. Now change the code
to call `add_one()` 100 times. Hopefully you can see that this is really
starting to help now!

And you can probably imagine that if it was any more complicated that just
copying and pasting one line it would make it much easier to see that it was
doing exactly the same thing each time too.

Finally, add in the single line `print(paste("Loop number", loop))` into the
loop at the end (between lines 11 and 12) and run it. You'll now be able see
each step of the loop running. We'll use techniques like this later to check
that the loop is doing the right thing.

*Note also that unlike functions, variables changed in `for` loops do change
what happens outside the loop, so `current.value` is `100` at the end, and not
still `0`.*

There is a section in R4DS (within a more general chapter on
[*Iteration*][R4DS-iteration]) that covers [`for` loops][for-R4DS]. For loops
are also covered by R Coder [here][for-rcoder].

### `while` loops

`while` loops allow us to repeat something **while** a condition is still `TRUE`.
They are particularly useful when you don't know in advance exactly how many
times something will be done, but can be used in any situation. Unlike `for`
loops though, you need to update everything manually each time around:

```{r whileloop, exercise = TRUE}
x <- 1
test <- x < 3 
while (test) {
  # Do something
  print(x)
  # Update x and update test
  x <- x + 1
  test <- x < 3 
}
```

`test` is checked at the beginning of each iteration, and as long as `test` 
continues to be
`TRUE`, the code block inside the curly brackets `{ ... }` is run. Be careful...
if you don't update `test` inside the curly brackets, your loop will continue 
forever! However, if (when!) `test` is finally `FALSE`, the loop will end and R
will continue on to the next line of code after the loop.


There is a
section in R4DS (within a more general chapter on [*Iteration*][R4DS-iteration])
that covers [`while` loops][while-R4DS]. While loops are also covered by R Coder
[here][while-rcoder].

[R4DS-iteration]: https://r4ds.had.co.nz/iteration.html
[for-R4DS]: https://r4ds.had.co.nz/iteration.html#for-loops
[while-R4DS]: https://r4ds.had.co.nz/iteration.html#unknown-sequence-length
[for-rcoder]: https://r-coder.com/for-loop-r
[while-rcoder]: https://r-coder.com/while-loop-r

## Exercises

Now try the following exercise, combining an `if` statement with a `for` loop to
print greetings (not `duck`):

```{r forexercise, exercise = TRUE}
greetings <- c("Hello", "Goodbye", "duck")
for (word ___) {
  if (___) {
    print(___)
  } 
}
```

```{r forexercise-solution}
greetings <- c("Hello", "Goodbye", "duck")
for (word in greetings) {
  if (word != "duck") {
    print(word)
  } 
}
```

```{r forexercise-check}
grade_this_code()
```

Solve the following problem -- a countdown from 10 to 1. 

```{r whileexercise, exercise = TRUE}
x <- 10
while (___) {
  # Do something
  print(x)
  if (x == 1) print("Blast off!")
  
  # Update test
  x <- ___
}
```

```{r whileexercise-solution}
x <- 10
while (x > 0) {
  # Do something
  print(x)
  if (x == 1) print("Blast off!")
  
  # Update test
  x <- x - 1
}
```

```{r whileexercise-check}
grade_this_code()
```
