---
title: "Project 3: Shannon entropy and the Berger-Parker index"
output:
  learnr::tutorial:
    theme: lumen
    includes:
      in_header: !expr system.file("tutorials/google-font.html", package = "RPiR")
runtime: shiny_prerendered
description: >
  Final project, adding in code to cope with Shannon entropy and Berger-Parker
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.reveal_solution = FALSE)
gradethis::gradethis_setup()

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```


## Overview

We are now going to look at two special cases, where the equation we used 
before needs to be adapted -- Shannon entropy, which is $\log(D_1)$, and the
Berger-Parker index, which is $\frac{1}{D_\infty}$, and adapt the code to cope
with these.

$$D_1 = \prod_{i \in \{1 \dots N_S, p_i \neq 0\}} p_i^{-p_i}$$

$$D_\infty = \frac{1}{\max_i p_i}$$

### Background

**Shannon entropy** is an extremely well-known mathematical measure of the
information content of a set of data. It is used here as a measure of 
"surprise" -- how surprising the species of the next individual observed in a 
sequence is likely to be. Where there is only 1 species, each observation is 
the same, and there is no surprise. However, when there are lots of species 
that are all equally abundant (we are ignoring how well mixed the population 
is), every encounter is likely to surprise us.

The **Berger-Parker** index simply tells us how dominant the most abundant 
species is. It ignores everything else. The lower the index, the less dominant 
the most abundant species is, and so the more other species are also observed.

## Tasks

You may want to copy your existing function code from Project 2 as a starting 
point for this exercise. If you do, remember that you need different function 
names in this exercise so that they don't conflict. 

You will be writing a function to calculate diversity, and then two functions to
calculate species richness and Simpson diversity from it as you did in Project
2. Then you need to write two new very simple functions which calculate the
Shannon entropy and the Berger-Parker index, using the generic function you
created above and **not** by calculating them from scratch (they will also need
the small amount of additional calculation -- **Shannon entropy is actually 
$\log(D_1)$** and **Berger-Parker is $\frac{1}{D_\infty}$** -- as detailed above
in the overview). You will probably find that the functions as written do not 
work! You may choose to test what the results should be again by comparing the 
outputs from the generic function at $q=1$ and $q=\infty$ (infinity, written as 
`Inf` in R) to matching results from the `rdiversity` package.

You should see that the generic function from Project 2 produces very different 
answers for a value of $q = 1$ and a value near 1 (say $q = 0.99$ or 
$q = 1.01$), and similarly for $q = \infty$ and a reasonably large number (say 
$q = 20$ or $q = 30$). In fact, you should actually get errors 
for exactly 1 and $\infty$, though you may in fact just produce the wrong 
answers (that are inconsistent with "nearby" numbers). In the former case, this 
is because we calculate $\frac{1}{0}$ in the equation (which is not a number), 
and in the latter we 
have a rounding error and end up calculating $0^0$ instead of 
$0.000001^{0.000001}$ which turn out to be very different 
numbers.

You will need to rewrite the general function using: 

```{r, eval = FALSE}
if (some_test) {
  # do something
} else if (some_other_test) {
  # do something different
} else {
  # a third thing
}
``` 

So that it checks for these two specific values of $q$, and instead use 
the following two equations: 

$$D_1 = \exp(-\sum(p \times \log(p)))$$
which is to say `exp(-sum(p * log(p)))` in R, where `p` is the vector of non-zero
species proportions.

$$D_\infty = \frac{1}{\max(p)}$$
which is just `1/max(p)` in R. You should to end up writing the updated general
function and the four specific functions that call this new updated function to
calculate the four measures of diversity we have discussed so far.

### Running the code

Run the two new functions again to make sure there are no errors, and also 
make sure that the generic function now gives the same (or very similar) 
results for 1 and 1.01, and for 20 and $\infty$.

### Testing for equality to exact values
**NB** It is just possible that you may find a problem with the code still not
giving you the correct answer when $q=1$. You can test for $q$ being infinity by
just by comparing the value to `Inf`, but this doesn't work reliably for other
numbers -- what goes wrong is a rounding error that is very common in computer 
programming. Whole numbers cannot be correctly represented as decimals without 
small errors creeping in -- so 1 may actually appear as 1.000000000001 -- and 
testing whether it is then equal to 1 will fail. Instead you should test 
whether it is sufficiently close to 1 for your purposes. Fortunately, we can do 
this by saying `if (abs(q-1) < 0.000001)`, say, instead of `if (q==1)`, which 
will check whether you are within 0.000001 of 1 -- *i.e.* 0.999999 to 1.000001. 
I've picked 0.000001, but you can pick another value if it seems better. R 
actually provides a number for this itself called the "machine 
epsilon", `.Machine$double.eps`, which is about $10^{-16}$, but it may 
be too small for our purposes. People often use `sqrt(.Machine$double.eps)` as 
a compromise (about $10^{-8}$). I honestly don't know why except that it looks 
quite technical, but feel free to use it if you like!

## Demo
Now write a demo that loads in this new functionality and then calculates the 
four diversity measures we have discussed for each dataset.

The demo should then call the general function directly (not the 4 specific 
functions) and just calculate the general diversity measures for the datasets 
provided with $q = 0$, 1, 2, and $\infty$ (remember, these correspond to the 4 
individual diversity measures we have investigated so far). Do you see a 
progression in the values you observe within each set of population counts as 
$q$ increases? Comment on this in the report.
