---
title: "Project 0: title"
output:
  learnr::tutorial:
    theme: lumen
    includes:
      in_header: !expr system.file("tutorials/google-font.html", package = "RPiR")
runtime: shiny_prerendered
description: >
  Final project.
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.reveal_solution = FALSE)
gradethis::gradethis_setup()

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```


## Overview

*In this series of practicals you will develop the code to calculate diversity
measures. In the first practical we will calculate the species richness and 
Simpson diversity of a population.*

## Background

R provides us with many built-in datasets, but sadly none of them are counts of
species abundances, so you will use the famous Barro Colorado Island dataset, a 
30+ year study of a 50-hectare forest plot in Panama. We will use the number of 
trees in the study site in 2010, which you have access to from the package you 
have developed during the course.

You will also construct a couple of fake datasets that you will use for 
comparison. Calculating the diversity of these populations (remembering that 
there are many, in fact a continuum of, diversity measures) will tell us how 
many observed species there are – the species richness – and give us measures 
of how evenly distributed the trees are amongst the tree species (the evenness 
of the species distribution) by various measures. **Species richness** is a 
very commonly used measure of diversity, which simply **counts the number of
non-zero species abundances**; **Simpson diversity** is also common: it simply
**calculates the species proportions from the counts (so the total sums to 1),
square them all individually, and add them all together** – it tells us (amongst 
other things) how likely two randomly chosen individuals are to be from the same 
species.

## Tasks

We are taking the data from the 2010 census – the true abundances for the whole 
site can be calculated as follows:

```{r, eval = FALSE}
library(BCI.package.name) 
tree.pop <- rowSums(bci_2010)
```

The total counts of trees of different species in this forest plot are now 
available in the variable tree.pop. We can then do some simple extraction of
information from the full species counts:

```{r, eval = FALSE}
species.names <- names(tree.pop) 
num.sp <- length(tree.pop) 
num.trees <- sum(tree.pop) 
mean.trees <- mean(tree.pop)
```

The next two lines sample a single random quadrat and sums up 10 random 
quadrats, and store the results in `quadrat.pop` and `quadrat10.pop`,
respectively:

```{r, eval = FALSE}
quadrat.pop <- rowSums(sample_by_subcommunities(bci_2010, 1)) 
quadrat10.pop <- rowSums(sample_by_subcommunities(bci_2010, 10))
```

You will also use a handful of fake datasets for comparison:

```{r, eval = FALSE}
one.pop <- c(num.trees, rep(0, num.sp - 1)) 
uneven.pop <- c(num.trees / 2, rep(0, num.sp - 1)) + rep(mean.trees / 2, num.sp) 
mixed.pop <- (1:num.sp) / sum(1:num.sp) * num.trees
even.pop <- rep(mean.trees, num.sp)
rand.pop <- as.vector(rmultinom(1, num.trees, rep(1 / num.sp, num.sp)))
rand50.pop <- as.vector(rmultinom(1, round(num.trees / 50), rep(1 / num.sp, num.sp)))
```

Remember that `rep(x, size)` repeats `x`, `size` times to make a vector of
length `size`, and `rmultinom()` draws from the **multinomial distribution**, in 
this case the first call itself creates a population of 222,718 individuals 
drawn from 301 species, each equally abundant (1/301).

Create a project (or a package), and in one file, write and document one
function that takes a vector of abundances like the data above as an argument 
and calculates the species richness from it. Hint: Applying a test to a vector 
of numbers, will apply the test to all of the numbers in the vector and produce 
a vector of Booleans (`TRUE` or `FALSE`) values, so `c(1, 3, 5, 7) > 4` returns 
`c(FALSE, FALSE, TRUE, TRUE)`. And `sum()` will count the number of `TRUE`s in 
a such vector of booleans. If you think about it, you should be able to see how 
to use these two ideas to count the number of non-zero abundances in a 
population.

Then write and document a second function that will calculate the Simpson 
diversity of the population, remembering, first, that this measure works with 
the relative proportions of the species not the total counts and, second, that 
we can do calculations with the whole vector at once.

## Running the code

Look at the values in the nine population vectors – you could try sorting them 
using `sort(x.pop)` and plotting them using `plot(sort(x.pop))`, 
`barplot(sort(x.pop))` (or something more attractive?) to get a clearer idea of 
what they look like – and decide which you think have the richest species 
diversity.

Run the two functions on the populations provided as see whether the results of 
the two diversity measures agree with your assessments. Note that a lower value 
means a higher diversity for the Simpson diversity measure.

## Report

In the project or package that you have created, and write a script that loads 
in the functions (either using `source()` on the file if you are using a 
project, or by building the package and calling `library()` on the package in
other case). The script then needs to read in the BCI data and in the rest of 
it you should generate a report that shows some kind of plot of each of the 
series of counts, and then calculates their species richness and Simpson 
diversity. If your function code is also be documented using the notebook 
report format, build a report for that file too. If it is a package, I should 
be able to use `?species_richness` or whatever you have called the functions to 
read the help files.

Remember to commit the code once you are happy before moving on to the next 
exercise – you can always come back later and make further commits if you 
realise you need to change something. I won't remind you about this in 
subsequent exercises.

## Note on function names

Note that you will often need to refer back to functions in later exercises, as 
you did occasionally in the practicals, and you will be writing several very 
similar functions in successive exercises that you might be tempted to give the 
same names. You will need to make sure they do not have identical names though 
or you won't be able to call them all. The simplest way of doing this might be 
to number them and call them `species_richness_1`, `species_richness_2`, *etc.* 
depending on what exercise you are in, but you are welcome to find a more 
elegant solution – anything that distinguishes them is fine.
