---
title: "Reproducible Programming in R Practical 4-3"
output: 
  learnr::tutorial:
    theme: lumen
    includes:
      in_header: !expr system.file("tutorials/google-font.html", package = "RPiR")
runtime: shiny_prerendered
description: >
  Students add functions and documentation to a data package.
---

```{r setup, include=FALSE}
library(learnr)

knitr::opts_chunk$set(error = TRUE)
set.seed(123)
```


## Overview
*In this session you will add some functions to your biodiversity data package* 
*to allow you to subsample from your data. The intention is just to see how we* 
*add functions to a package, and to provide you with an example of what the*
*documentation for a function in an R package looks like. You will also add the*
*documentation for the package as a whole.*

## Background

It’s rarely the case that you will have access to a fully sampled dataset like
that at BCI. To pretend that your dataset has incomplete data, I have provided 
you with two functions on Moodle that will subsample the data in different ways. 
The first, `sample_by_species()`, will sample from the dataset as if you were
recording only some of the species (chosen at random) and ignoring the others. 
The second, `sample_by_subcommunities()`, will record the counts from only 
some of the subcommunities (here quadrats) and ignore the others. There are 
other ways of subsampling or recording data incompletely, and we will
investigate this in the project. You can see some documentation on how to write
help files for functions at https://r-pkgs.org/man.html#man-functions and an
older version at http://r-pkgs.had.co.nz/man.html#man-functions – these pages 
are very useful and you should take a look.

## Tasks

You simply need to download the R files for these functions from Moodle and put
them in the <span style="color: #de77ae;">R</span> folder of your package, and
edit the documentation for each of these functions and the 
<span style="color: #de77ae;">DESCRIPTION</span> file of your package so that 
all dependencies are included. Then regenerate the documentation and reinstall 
the package from the <span style="color: #de77ae;">Build</span> menu. Doing this 
should allow you to access the help files for these functions 
(`?sample_by_species` and `?sample_by_communities`) and run the example code (by 
calling `example(sample_by_species)`, for instance). 

You should also create a description for the overall package, so that `?BCI` 
returns a description, and you should put the same (or similar) text into the 
<span style="color: #de77ae;">README.md</span> file for the package so that 
people going to GitHub will see what the package that you have created does 
without having to install it. Note that GitHub README pages use a special type 
of markdown called Git Flavored Markdown; more information can be found here https://docs.github.com/en/free-pro-team@latest/github/writing-on-github with 
more general documentation here 
https://guides.github.com/features/mastering-markdown/. Note that with markdown
you don’t need to start lines with `#'`.

To do this, create a new file in the <span style="color: #de77ae;">R</span> 
folder – I’m going to call it <span style="color: #de77ae;">BCI-package.R</span> 
because the convention is {packagename}-package.R. Then describe your package (most easily lifted 
from the <span style="color: #de77ae;">DESCRIPTION</span> file):

```{r eval = FALSE}
#' BCI data package 
#'
#' Package to hold the BCI data (or whatever) – maybe also mention something
#' about these functions now, and put that in the DESCRIPTION too. And then
#' put it in the README.md file. 
#'
#' @import magrittr 
#'
#' @name BCI-package
#' @aliases BCI
#' @docType package 
#'
NULL
```

There are a few things going on here that you should notice for making packages 
in the future. The first is that you need to say what your package is called – 
here I have given it two names `BCI-package` and an alias of just `BCI`, which 
you do with the `@aliases` command. Secondly, the `NULL` at the end of the file
is included when there is no object associated with this documentation, which is
the case here, since this file contains the package documentation, as defined
explicitly in the `@docType` tag (more info here –
http://r-pkgs.had.co.nz/man.html#man-packages). Finally, if you are using any 
packages in any of your functions, you may want to import them into your 
package. This is done using `@import` – see more here –
http://r-pkgs.had.co.nz/namespace.html. You can then put the same, or a similar,
package description into the <span style="color: #de77ae;">README.md</span> 
file. Once you have redocumented and reinstalled everything, and checked that it 
is all working, commit and push all of the changes to GitHub and ask your 
collaborator to check that this works for them as well.
